---
title: "Subset the trait database"
author: "Patrick Pata"
date: "2023-10-27"
output: html_document
---

This file subsets the zooplankton trait database for relevant traits required in the analysis of the associated manuscript. There is no need to rerun the scripts within this file since the output data is already produced as saved in the data/ folder. If the curation of the trait database changes, please uncomment the data saving scripts in lines 270 and 498. 

This will be based on Level 1 trait data. New information can be added to this without having to overhaul the trait database.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load libraries, functions, and data
```{r}
library(tidyverse)
library(here)
library(scales)
library(cowplot)
library(arsenal) # for the summary table

set_here(path = "..")
`%notin%` <- Negate(`%in%`)
theme_set(theme_bw())


mg.exclude <- c("Cladoceran","Cubozoan","Scyphomedusae","Decapod")


# * FUNCTIONS * 
# source(here("functions/toolkit_database.R"))
standard_temp <- function(rate0, t0, t = 15, Q10 = 2.8) {
  10^(log10(rate0) + log10(Q10)*((t-t0)/10) )
}
# Calculate the weighted means and SDs of traits when the dataset is a mix of 
#   individual values and averaged records.
getSpeciesMeanSD <- function(traitValue, dispersionSD, individualCount){
  # extract indiv values and grouped values
  val.indiv <- traitValue[which(individualCount == 1)]
  val.group.mean <- traitValue[which(individualCount > 1)]
  val.group.N <- individualCount[which(individualCount > 1)]
  val.group.sd <- dispersionSD[which(individualCount > 1)]
  # calculate mean of the individual values
  val.indiv.mean <- mean(val.indiv)
  val.indiv.N <- length(val.indiv)
  val.indiv.sd <- sd(val.indiv, na.rm = TRUE)
  
  # calculate N
  N <- sum(individualCount)
  # calculate weighted mean
  mean <- weighted.mean(c(val.indiv.mean, val.group.mean),
                        c(val.indiv.N, val.group.N))
  # calculate pooled sd
  if (val.indiv.N == 1) {
    sd <- pooled.sd(c(val.group.sd),
                    c(val.group.N))
  } else {
    sd <- pooled.sd(c(val.indiv.sd, val.group.sd),
                    c(val.indiv.N, val.group.N))
  }
  # calculate standard error
  se <- sd / sqrt(N)
  
  return(list("N" = N, "mean" = mean, "sd" = sd, "se" = se))
}

# pooled sd
pooled.sd <- function(SDarray, Narray){
  sqrt(sum((Narray-1)*(SDarray)^2) / (sum(Narray) - length(Narray)) )
  # # if inputs are: x,y,Nx,Ny
  # sqrt( ((Nx-1)*x^2 + (Ny-1)*y^2) / (Nx+Ny-2))
}

cleanStrings <- function(A){
  A <- str_split(A, pattern =";", simplify = TRUE) 
  A <- str_trim(A)
  A <- sort(A)
  A <- str_replace(A, "NA", replacement = "")
  A <- A[A != '']
  A <- unique(as.list(A))
  A <- paste(A, collapse = "; ")
  A <- as.character(A)
}
```

# 2. Prepare data
```{r, eval=TRUE}
# Taxonomy table
taxonomy <- read.csv(here("data/trait_database_raw/taxonomy_table_20230628.csv")) %>% 
  distinct(taxonID, .keep_all = TRUE)

trait.list <- c("bodyLength","carbonWeight","dryWeight","wetWeight",
                "waterPWW","carbonPDW","nitrogenPDW","phosphorusPDW",
                "nitrogenTotal","phosphorusTotal",
                "ratioCN","ratioCP","ratioNP",
                "respirationRate_15C","respirationRate_WSC_15C",
                "excretionRateN_15C","excretionRateN_WSDW_15C",
                "excretionRateP_15C","excretionRateP_WSDW_15C",
                "clearanceRate_15C","clearanceRate_WSC_15C",
                "feedingMode","reproductionMode","trophicGroup")

# Original level 1 
lvl1 <- read.csv(here("data/trait_database_raw/trait_dataset_level1-2023-08-15.csv")) %>% 
  filter(traitName %in% trait.list) %>% 
  filter(valueType == "numeric") %>% 
  mutate(traitValue = as.numeric(traitValue)) %>% 
  # filter(majorgroup %notin% mg.exclude) %>% 
  
  # ** Manual curation **
  
  # Remove the records of Gigantocypris muelleri because this is the only species with phosphate composition for ostracods which is anomalously high in the analysis and the nitrogen composition is also anomalously high compared to the other ostracods - making estimates of stoichiometry and excretion rates for this species not representative of ostracoda in this analysis.
  filter(scientificName != "Gigantocypris muelleri") %>% 
  # Exclude the Clio pyramidata excretion rates which were observed for Clio sulcuta and measured at <0 degC in the Antarctic resulting in  high values when standardized to 8.5degC. The total N and P values are also very high.
  filter(!(scientificName == "Clio pyramidata" & 
             traitName %in% c("excretionRateN_15C","excretionRateN_WSDW_15C",
                              "excretionRateP_15C","excretionRateP_WSDW_15C",
                              "nitrogenTotal", "phosphorusTotal"))) %>% 
  # Exclude the carbon weight of "Euphilomedes producta" because it is 4 orders of magnitude  smaller compared to the carbon weight of Euphilomedes interpuncta which has more trait data. This results in an erroneous calculated WSC respiration rate that is 3 orders of magnitude greater than any organism.
  filter(!(scientificName == "Euphilomedes producta" & traitName == "carbonWeight")) %>% 
  filter(!(scientificName == "Mysida" & traitName == "carbonWeight")) %>% # very low carbon weights
  filter(!(scientificName == "Clytia" & traitName == "nitrogenPDW")) %>% # value too high when most species have low values
  filter(!(scientificName == "Rathkea octopunctata" & traitName == "nitrogenPDW" 
           & primaryReference == "Matsakis1999")) %>% 
  filter(!(scientificName == "Acartia" & traitName == "phosphorusPDW")) %>% 
  # exclude some very high C:N records which are beyond the range for similar taxa
  filter(catalogNumber %notin% c("38-5223-4","38-2654-1",
                                 "38-630-14","38-630-31")) %>% 
  # exclude clearance rates for appendicularians, will be added back later
  filter(!(majorgroup == "Appendicularian" & 
           traitName %in% c("clearanceRate_15C"))) %>% 
  # exclude clearance rates for Tortanus (Boreotortanus) discaudatus which is a magnitude higher than other Tortanus values
  filter(!(scientificName == "Tortanus (Boreotortanus) discaudatus" &
             traitName %in% c("clearanceRate_15C","clearanceRate_WSC_15C"))) 
# convert dryWeightPWW to waterPWW
lvl1.dwww <- read.csv(here("data/trait_database_raw/trait_dataset_level1-2023-08-15.csv")) %>% 
  filter(traitName == "dryWeightPWW") %>% 
  mutate(traitValue = as.numeric(traitValue)) %>% 
  mutate(traitName = "waterPWW", traitID = 19,
         traitValue = 100-traitValue,
         basisOfRecord = "derived from dryWeightPWW")

# Original level 1 excluded
lvl1.ex <- read.csv(here("data/trait_database_raw/trait_dataset_level1_excluded-2023-08-15.csv")) %>% 
  filter(valueType == "numeric") %>% 
  filter(traitName %in% trait.list) %>% 
  # filter(traitName %in% c())
  filter(traitName != "bodyLength") %>% 
  filter(stageID %notin% c(23:37,46)) %>% 
  filter(lifeStage %notin% c("larvae","calyptopsis","furcilia","egg",
                             "trophozooid","juvenile")) %>% 
  mutate(traitValue = as.numeric(traitValue)) %>% 
    # anomalous values
    filter(catalogNumber %notin% c("16-3559-1","23-630-53","28-3025-63",
                                   "31-637-22","31-1423-6")) %>% 
  filter(exclusionType != "duplicate")


# *** New data ***

# Updated clearance rates for appendicularians + data on thaliaceans and thecosomes
new.clearance <- read.csv(here("data/new_trait_info/level1_appendicularian_clearance_rate_updated.csv")) %>% 
  select(-c(taxonID)) %>% 
  left_join(distinct(taxonomy,scientificName, taxonID, phylum, class, order, 
                     family, genus, acceptedNameUsageID, acceptedNameUsage,
                     taxonRank, taxonRankID, aphiaID),
            by = "scientificName") %>% 
  # Standardize to 15 degC
  mutate(traitValue = standard_temp(traitValue, assocTemperature,
                                    t = 15, Q10 = 2.8),
         traitName = paste0(traitName,"_15C"),
         assocTemperature = "15", 
         verbatimTemperature = as.character(verbatimTemperature)) 

# New composition data 
new.composition <- read.csv(here("data/new_trait_info/Gorsky1988a.csv")) %>% 
  left_join(distinct(taxonomy,taxonID, taxonID, phylum, class, order, 
                     family, genus, acceptedNameUsage,
                     taxonRank, taxonRankID),
            by = "taxonID") %>% 
  mutate(verbatimTraitValue = as.character(verbatimTraitValue))

# The TorresBailey2022 Respiration rates as weight-specific WW
new.respWW <- read.csv(here("data/new_trait_info/TorresBailey2022_RespirationRates.csv")) %>% 
  filter(!is.na(sizeAssocValue)) %>% 
  # convert size from g to mg WW
  mutate(sizeAssocValue = sizeAssocValue * 1000,
         sizeAssocUnit = "mg") %>% 
  # convert rate from wet-weight specific to individual rate
  mutate(traitValue = traitValue * sizeAssocValue,
         traitUnit = "ul O2 h^-1",
         traitName = "respirationRate") %>% 
  # standardize to 8.5 degC
  mutate(traitValue = standard_temp(traitValue, t0 = traitValueTemperature, 
                                    t = 8.5, Q10 = 2.8),
         traitValueTemperature = 8.5) %>% 
  # exclude very smal values likely because wet weight is wrong
  filter(traitValue > 0.05) %>% 
  mutate(verbatimTraitValue = as.character(verbatimTraitValue),
         verbatimTemperature = as.character(verbatimTemperature))

# *** Combine all data and export ***
traits.lvl1 <- bind_rows(lvl1, lvl1.ex, lvl1.dwww,
                         new.clearance, new.composition, new.respWW) %>% 
  select(catalogNumber, taxonID, scientificName, traitName, traitValue,
         traitUnit, lifeStage, assocTemperature, individualCount, dispersionSD,
         sizeAssocName, sizeAssocUnit, sizeAssocValue, sizeType,
         primaryReference, secondaryReference, basisOfRecord,
         taxonRank, phylum, class, order, family, genus, majorgroup) %>% 
  filter(majorgroup %notin% c("Cubozoan","Cladoceran")) %>%
  
  
  # Update rates to standard temperature of 8.5 degC
  mutate(assocTemperature = if_else(grepl("_15C", traitName),
                                    "8.5",assocTemperature)) %>% 
  mutate(traitName = str_remove(traitName,"_15C")) %>% 
  mutate(traitValue = if_else(grepl("respiration|clearance",traitName),
                              standard_temp(traitValue, t0 = 15, 
                                            t = 8.5, Q10 = 2.8),
                              traitValue)) %>% 
  mutate(traitValue = if_else(grepl("excretionRateN",traitName),
                              standard_temp(traitValue, t0 = 15, 
                                            t = 8.5, Q10 = 2.0),
                              traitValue)) %>% 
  mutate(traitValue = if_else(grepl("excretionRateP",traitName),
                              standard_temp(traitValue, t0 = 15, 
                                            t = 8.5, Q10 = 1.55),
                              traitValue)) %>% 
  

  # rename nitrogen and phosphorus weight
  mutate(traitName = str_replace(traitName,"Total","Weight")) %>% 
  # Set factor levels
  mutate(majorgroup = 
           factor(majorgroup, 
                  levels = c("Appendicularian","Non-calanoid","Calanoid",
                             "Ostracod","Mysid","Amphipod", "Euphausiid",
                             "Polychaete","Pteropod","Chaetognath",
                             "Hydromedusae","Siphonophore","Thaliacean",
                             "Ctenophore","Scyphomedusae","Decapod")),
         traitName = 
           factor(traitName, 
                  levels = c("bodyLength","dryWeight","wetWeight","waterPWW",
                             "carbonWeight","nitrogenWeight","phosphorusWeight",
                             "carbonPDW","nitrogenPDW","phosphorusPDW",
                             "ratioCN","ratioCP","ratioNP",
                             "respirationRate","respirationRate_WSC",
                             "excretionRateN","excretionRateN_WSDW",
                             "excretionRateP","excretionRateP_WSDW",
                             "clearanceRate","clearanceRate_WSC")))


trait.list <- str_remove(trait.list,"_15C")

majorgroup.colors <- data.frame(
  majorgroup = c("Appendicularian","Non-calanoid","Calanoid","Ostracod",
                 "Mysid","Amphipod", "Euphausiid", "Polychaete","Pteropod",
                 "Chaetognath", "Hydromedusae","Siphonophore",
                 "Thaliacean","Ctenophore","Scyphomedusae","Decapod"),
  color = c("violetred","dodgerblue","blue3","forestgreen",
                       "seagreen3","slateblue1","cyan3","lightpink2","green2",
                       "darkorchid4", "tan1","yellow2",
                       "tomato1","red3","orange4","salmon3"))
                       


# rm(lvl1, lvl1.ex, new.clearance, new.composition, lvl1.dwww)

# save(traits.lvl1, taxonomy, trait.list, majorgroup.colors,
#      file = here("data/traits_level1_curated_20231027.RData"))
```


# 3. Create major group summary table and figures
```{r, fig.width=6}
# Summary table
mg.table <- traits.lvl1 %>% 
  group_by(traitName, majorgroup) %>% 
  summarise(nobs = n(), nsp = length(unique(scientificName)),
            tv.mean = mean(traitValue), tv.sd = sd(traitValue),
            tv.min = min(traitValue), tv.max = max(traitValue),
            .groups = "drop")

# 
# my_controls <- tableby.control(
#   test = T,
#   total = T,
#   numeric.stats = c("N","meansd","range"),
#   stats.labels = list(
#     meansd = "Mean (SD)",
#     range = "Min - Max",
#     N = "Nrecords"),
#   digits = 2
# )
# 
# traits.lvl1.long <- traits.lvl1 %>% 
#   select(majorgroup, traitName, traitValue) %>% 
#   pivot_wider(names_from = traitName, values_from = traitValue) 
#   
# mg.table <- tableby(majorgroup ~ traitValue,
#                     data = traits.lvl1 %>% filter(traitName == "carbonPDW"),
#                     control = my_controls)
# summary(mg.table)

# Number of records and species
ggplot(mg.table, aes(x = majorgroup, y = traitName)) +
  geom_tile(aes(fill = nobs)) +
  geom_text(aes(label = nsp), color = "black",) +
  scale_y_discrete(limits = rev)+
  scale_fill_distiller(limits = c(0,200), oob = scales::squish) +
  scale_x_discrete(guide = guide_axis(n.dodge = 2)) 


# ggsave(filename = here("figures/trait_distribution_numspecies.png"),
#     width = 10, height= 8, units = "in", dpi = 300, bg = "white")


# Distribution of data
g1 <- ggplot(traits.lvl1 %>% filter(grepl("PDW|PWW|CP|NP|CN", traitName)), 
       aes(x = majorgroup, y = traitValue, color = majorgroup)) +
  geom_boxplot() +
  scale_color_manual(values = majorgroup.colors$color, name = "major group") + 
  theme(axis.text.x = element_blank(), axis.title.x = element_blank(),
        legend.position = "none") +
  facet_wrap(~traitName, scales = "free_y", ncol = 4)

g2 <- ggplot(traits.lvl1 %>% filter(!grepl("PDW|PWW|CP|NP|CN", traitName)), 
       aes(x = majorgroup, y = traitValue, color = majorgroup)) +
  geom_boxplot() +
  scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
                labels = trans_format("log10", math_format(10^.x))) +
  scale_color_manual(values = majorgroup.colors$color) + 
  theme(axis.text.x = element_blank(),
        legend.position = "none") +
  facet_wrap(~traitName, scales = "free_y")

legend_b1 <- get_legend( g1 + 
                           guides(color = guide_legend(ncol = 1)) +
                           theme(legend.position = "right"))

plot_grid(plot_grid(g1, g2, ncol = 1, rel_heights = c(2,4)),
          legend_b1, nrow = 1, rel_widths = c(10,1))

# ggsave(filename = here("figures/extra_trait_distribution_level1.png"),
#     width = 15, height= 10, units = "in", dpi = 300, bg = "white")
```


# 4. Species level data
## Calculate species level averages
These are weighted means when the N and sd were reported.
```{r}
# # The curated level 1 data based on the chunk above
# load(here("data/traits_level1_curated_20231027.RData"))

# Curate the body length data to use maximum length for copepods and for females when possible.
A <- traits.lvl1 %>% 
  filter(traitName == "bodyLength")

# For copepods, select females and total length when possible
A.1 <- A %>% 
  filter(class == "Copepoda") %>% 
  filter(lifeStage == "female" & sizeType == "total length")
# females but not in total length
A.2 <- A %>% 
  filter(class == "Copepoda") %>% 
  filter(taxonID %notin% A.1$taxonID) %>% 
  filter(lifeStage == "female")

# then select "adult"
A.3 <- A %>% 
  filter(taxonID %notin% A.1$taxonID) %>% 
  filter(taxonID %notin% A.2$taxonID) %>% 
  filter(lifeStage %in% c("adult"))

# Get average for adults and females
A.4 <- bind_rows(A.1, A.2, A.3) 

# all taxa not in A.4
trait.length <- A %>% 
  filter(taxonID %notin% A.4$taxonID) %>% 
  bind_rows(A.4)

rm(A, A.1, A.2, A.3, A.4)


traits.species <- traits.lvl1 %>% 
  filter(traitName != "bodyLength") %>% 
  # return the curated length information
  bind_rows(trait.length) %>% 
  
  
  # if N is NA or if sd is not reported, N = 1
  mutate(individualCount = if_else(is.na(individualCount), 1, individualCount)) %>%
  mutate(individualCount = if_else(is.na(dispersionSD), 1, individualCount)) %>% 
  # calculate trait-taxon average
  group_by(traitName, traitUnit, taxonID, scientificName) %>%  
  # TODO: improve this mutate so it doesn't do the calculation thrice
  summarise(nRecords = n(),
         tvN = getSpeciesMeanSD(traitValue, dispersionSD, individualCount)$N,
         tvSD = getSpeciesMeanSD(traitValue, dispersionSD, individualCount)$sd,
         tvVal = getSpeciesMeanSD(traitValue, dispersionSD, individualCount)$mean,
         .groups = "drop") %>%
  mutate(traitValue = tvVal, individualCount = tvN, dispersionSD = tvSD,
         basisOfRecord = "calculated taxon average") %>% 
  dplyr::select(-c(tvVal, tvN, tvSD)) 
```


## Convert categorical traits to binary per species
 Work on level 1 data
```{r}
trait.cat <- read.csv(here("data/trait_database_raw/trait_dataset_level1-2023-08-15.csv")) %>% 
  filter(traitName %in% c("trophicGroup","feedingMode","reproductionMode")) %>% 
  # exclude some feeding modes that are not specified
  filter(traitValue %notin% c("ambush","active","passive")) %>% 
  group_by(traitID, taxonID) %>% 
  mutate(traitValue = cleanStrings( paste(traitValue, collapse = "; ")) )%>% 
  ungroup() 


# Trophic group
trait.TG <- trait.cat %>% 
  filter(traitName == "trophicGroup") %>% 
  mutate(TG.carnivore = if_else(grepl("carnivore",traitValue)==TRUE,1,0)) %>% 
  mutate(TG.omnivore = if_else(grepl("omnivore",traitValue)==TRUE,1,0)) %>% 
  mutate(TG.herbivore = if_else(grepl("herbivore",traitValue)==TRUE,1,0)) %>% 
  mutate(TG.detritivore = if_else(grepl("detritivore",traitValue)==TRUE,1,0)) %>% 
  select(-c(traitName, traitValue)) %>% 
  pivot_longer(cols = c(TG.carnivore, TG.omnivore, TG.herbivore, TG.detritivore), 
               names_to = "traitName", values_to = "traitValue") %>% 
  mutate(valueType = "binary") %>% 
  filter(traitValue == 1) %>% 
  relocate(traitName, traitValue)

trait.FM <- trait.cat %>% 
  filter(traitName == "feedingMode") %>% 
  mutate(FM.cruise = if_else(grepl("cruise",traitValue)==TRUE,1,0)) %>% 
  mutate(FM.current = if_else(grepl("current",traitValue)==TRUE,1,0)) %>% 
  mutate(FM.active.ambush = if_else(grepl("active ambush",
                                       traitValue)==TRUE,1,0)) %>% 
  mutate(FM.passive.ambush = if_else(grepl("passive ambush",
                                        traitValue)==TRUE,1,0)) %>% 
  # mutate(FM.active = if_else(grepl("active",traitValue)==TRUE & 
  #                         grepl("active ambush",traitValue)==FALSE,
  #                         1,0)) %>%  
  # mutate(FM.passive = if_else(grepl("passive",traitValue)==TRUE,1,0)) %>% 
  # mutate(FM.ambush = if_else(grepl("ambush",traitValue)==TRUE,1,0)) %>% 
  # mutate(FM.particle.feeder = if_else(grepl("particle feeder",
  #                                        traitValue)==TRUE,1,0)) 
  select(-c(traitName, traitValue)) %>% 
  pivot_longer(cols = c(FM.cruise, FM.current, 
                        FM.active.ambush, FM.passive.ambush), 
               names_to = "traitName", values_to = "traitValue") %>% 
  mutate(valueType = "binary") %>% 
  filter(traitValue == 1) %>% 
  relocate(traitName, traitValue)

trait.RM <- trait.cat %>% 
  filter(traitName == "reproductionMode") %>% 
  mutate(RM.brooding = if_else(grepl("brood",traitValue)==TRUE,1,0)) %>% 
  mutate(RM.broadcasting = if_else(grepl("broadcast",traitValue)==TRUE,1,0)) %>% 
  select(-c(traitName, traitValue)) %>% 
  pivot_longer(cols = c(RM.brooding, RM.broadcasting), 
               names_to = "traitName", values_to = "traitValue") %>% 
  mutate(valueType = "binary") %>% 
  filter(traitValue == 1) %>% 
  relocate(traitName, traitValue)
 
traits.binary <- bind_rows(trait.RM, trait.FM, trait.TG) %>% 
  group_by(scientificName, traitName, traitValue, traitID, valueType,
           taxonID, taxonRank, phylum, class, order, family,
           genus, majorgroup, basisOfRecord) %>% 
  summarise(primaryReference = cleanStrings(paste(primaryReference, 
                                                  collapse = "; ")),
         secondaryReference = cleanStrings(paste(secondaryReference, 
                                                 collapse = "; ")),
         .groups = "drop") 

rm(trait.RM, trait.FM, trait.TG)
```

## Export species level data
```{r}
# Add taxonomy information
traits.species <- traits.species %>% 
  filter(!is.na(taxonID)) %>% 
  left_join(select(taxonomy, taxonID, taxonRank, phylum, class, order, family,
                   genus, majorgroup), 
            by = "taxonID") %>% 
  filter(majorgroup != "") %>% 
  mutate(valueType = "numeric")
traits.binary <- traits.binary %>% 
  mutate(valueType = "binary")

# save(traits.species, traits.binary,
#      file = here("data/traits_level2_curated_20231027.RData"))
```

