---
title: "Diversity and ecosystem functions"
author: "Patrick Pata"
date: "2024-11-12"
output: html_document
---

This file contains the exploratory analysis that relates the functional diversity indices with ecosystem functioning as proxied by the community level trait values.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load libraries and data
```{r}
library(tidyverse)
library(cowplot)
library(here)
library(ggtext)
library(vegan)
library(ggrepel)
theme_set(theme_bw())


# Load data
load(here("data/BC_bioregionalization_metadata.RData"))
load(here("data/fundiv_3721_163species_20240618.RData"))
load(here("data/community_metrics_3721_163species_20250217.RData"))

geomean <- function(x) {  exp(mean(log(x), na.rm = TRUE))  }
`%notin%` <- Negate(`%in%`)

fungrp.colors <- c("turquoise3", "steelblue2","darkorange2",
                   "orchid","magenta4","lightcoral",
                   "gold3","darkgreen","darkolivegreen4",
                   "olivedrab3")
                   
clrs.bioregion <- c("#004DFB","#FB20CA","#FFA500","#20FF00")

rm(zoop.163.ctotal)
```

# 2. Prepare the BEF matrix
```{r}
zoop.163.totals <- zoop.163.ctotal %>% 
  filter(Trait %in% c("totalCarbonWeight", "totalRespirationRate",
                      "totalExcretionRateN", "totalExcretionRateP",
                      "totalClearanceRate")) %>% 
  pivot_wider(names_from = "Trait", values_from = "CTotal") %>% 
  rename(totalCarbon = totalCarbonWeight)

B.mat <- fundiv.ind.biom %>% 
  rownames_to_column("Key")

EF.mat <- zoop.163.cwm %>% 
  filter(Trait %in% c("ratioCN")) %>% 
  pivot_wider(names_from = Trait, values_from = CWM) %>% 
  left_join(zoop.163.totals, by = join_by(Key, Bioregion))

BEF.mat <- left_join(B.mat, EF.mat,
                     by = join_by(Key, Bioregion))

# Subset the matrix for curve fitting
BEF.2.mod <- BEF.mat %>% 
  mutate(logRespiration = log10(totalRespirationRate),
         logExcretionN = log10(totalExcretionRateN)
         ) %>% 
  select(-c(totalCarbon:totalClearanceRate)) %>% 
  pivot_longer(cols = c(Sp.Ric:FOri),
               names_to = "Diversity", values_to = "Index") %>% 
  pivot_longer(cols = c(ratioCN:logExcretionN),
               names_to = "CTV", values_to = "Value") %>% 
  mutate(Diversity = factor(Diversity,
                            levels = c("Sp.Ric", "FRic", "FDis", 
                                       "FEve", "FDiv", "FOri")))
```


# 3. Curve fitting

Selection based on AIC. Using AIC reduces overfitting and is appropriate here because we are comparing between alternative models with different structures. The pseudo-R2 are also extracted because these are more familiar to interpret (only because it is scaled from 0 to 1). Generally, the relative differences between BEF pairs are not similar between these two metrics of model performance. Although highest pseudo R2 are from low AIC (better model).

## Setup the model fitting function
```{r}
# The general model fitting function 
fit_models <- function(data) {
  models <- list(
    linear = lm(Value ~ Index, data = data),
    # quadratic = lm(Value ~ Index + I(Index^2), data = data),
    logarithmic = tryCatch(
      nls(Value ~ a + b * log(Index), data = data, start = list(a = 1, b = 1)),
      error = function(e) NULL
    ),
    power = tryCatch(
      nls(Value ~ a * Index^b, data = data, start = list(a = 1, b = 1)),
      error = function(e) NULL
    ),
    saturating = tryCatch(
      nls(Value ~ a * Index / (b + Index), data = data, start = list(a = 1, b = 1)),
      error = function(e) NULL
    )
  )
  
  results <- lapply(names(models), function(model_name) {
    model <- models[[model_name]]
    if (is.null(model)) return(NULL)
    
    # Extract coefficients and standard errors
    if ("lm" %in% class(model)) {
      coef_summary <- summary(model)$coefficients
      stderr <- summary(model)$sigma
    } else {
      coef_summary <- summary(model)$parameters
      stderr <- sqrt(sum(resid(model)^2) / (nrow(data) - length(coef(model))))
    }
    
    data.frame(
      model_type = model_name,
      aic = AIC(model),
      stderr = stderr,
      # Pseudo R2 is residual sum of squares as proportion of total varation
      r2 = ifelse("lm" %in% class(model), summary(model)$r.squared,
                  1 - sum(resid(model)^2) / sum((data$Value - mean(data$Value))^2)),
      coefficients = paste(names(coef(model)), coef(model), 
                           sep = "=", collapse = "; "),
      standard_errors = paste(names(coef(model)), 
                              coef_summary[, "Std. Error"], 
                              sep = "=", collapse = "; ")
    )
  })
  
  do.call(rbind, results)
}

```


## Global analysis
```{r}
# Fit models for each combination of Diversity and CTV
results.global <- BEF.2.mod %>%
  group_by(Diversity, CTV) %>%
  summarise(
    model_summaries = list(fit_models(cur_data())),
    minX = min(Index),
    maxX = max(Index),
    .groups = "drop"
  ) %>%
  unnest(model_summaries) %>%
  group_by(Diversity, CTV) %>%
  slice_min(aic) %>%  # Select the model with the lowest AIC for each group
  rename(best_model = model_type)  # Add the best model name as a separate column

# # Determine instances when the r2 >= 0.1 
# # Log carbon and log respiration will produce similar curves so just focus on log carbon in the main figure and put log respiration in the supplementary figure.
# results.global.high <- results.global %>% 
#   filter(r2 >= 0.1)
```


## Bioregion-specific analysis
```{r}
# Fit models for each combination of Bioregion and CTV
results.bioregion <- BEF.2.mod %>%
  group_by(Bioregion, Diversity, CTV) %>%
  summarise(
    model_summaries = list(fit_models(cur_data())),
    minX = min(Index),
    maxX = max(Index),
    .groups = "drop"
  ) %>%
  unnest(model_summaries) %>%
  group_by(Bioregion, Diversity, CTV) %>%
  slice_min(aic) %>%  # Select the model with the lowest AIC for each group
  rename(best_model = model_type) %>%  # Add the best model name as a separate column
  ungroup()

# # Add a column for the global r2 and filter accordingly to only instances of R2 >= 0.1
# results.bioregion.high <- results.bioregion %>%
#   left_join(select(results.global, Diversity, CTV, r2.global = r2),
#             by = join_by(Diversity, CTV)) %>%
#   filter(r2.global >= 0.1)
```

# 4. Extra: Visual residual analysis
This is for visualizing all model types for one BEF combination. Change the diversity metric and CTV inside the filter() function. Residuals should be random around zero and for observed vs fitted, better if around the 1:1 line. 
```{r, eval=FALSE}
library(ggdist)

plot_data <- BEF.2.mod %>%
  filter(Diversity == "FRic", CTV == "logRespiration")
  
fit_results <- fit_models(plot_data)

models <- list(
  linear = lm(Value ~ Index, data = plot_data),
  logarithmic = tryCatch(
    nls(Value ~ a + b * log(Index), data = plot_data, start = list(a = 1, b = 1)),
    error = function(e) NULL
  ),
  power = tryCatch(
    nls(Value ~ a * Index^b, data = plot_data, start = list(a = 1, b = 1)),
    error = function(e) NULL
  ),
  saturating = tryCatch(
    nls(Value ~ a * Index / (b + Index), data = plot_data, start = list(a = 1, b = 1)),
    error = function(e) NULL
  )
)


plot_model_diagnostics <- function(data, model) {
  if (is.null(model)) {
    cat("Model is NULL, skipping diagnostics.\n")
    return(NULL)
  }
  
  # Extract fitted values and residuals
  fitted_values <- fitted(model)
  residuals <- resid(model)
  
  # Create a data frame for plotting
  diag_data <- data.frame(
    Observed = data$Value,
    Fitted = fitted_values,
    Residuals = residuals
  )
  
  # Plot 1: Residuals vs Fitted
  residual_plot <- ggplot(diag_data, aes(x = Fitted, y = Residuals)) +
    geom_point(color = "blue", alpha = 0.6) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    labs(title = "Residuals vs Fitted", x = "Fitted Values", y = "Residuals") +
    theme_minimal()
  
  # Plot 2: Fitted vs Observed
  fitted_vs_observed_plot <- ggplot(diag_data, aes(x = Observed, y = Fitted)) +
    geom_point(color = "green", alpha = 0.6) +
    geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed") +
    labs(title = "Fitted vs Observed", x = "Observed Values", y = "Fitted Values") +
    theme_minimal()
  
  # Plot 3: Density distribution of residuals
  residual_density_plot <- ggplot(diag_data, aes(x = Residuals)) +
    geom_density(color = "blue", fill = "lightblue", alpha = 0.5) +
    stat_halfeye(aes(y = 0), orientation = "horizontal") +
    labs(title = "Density Distribution of Residuals", x = "Residuals", y = "Density") +
    theme_minimal()
  
  # Combine all plots using cowplot
  combined_plot <- plot_grid(
    residual_plot, 
    fitted_vs_observed_plot, 
    residual_density_plot, 
    ncol = 2, 
    labels = c("A", "B", "C"),
    label_size = 14
  )
  
  # Display the combined plot
  print(combined_plot)
}


# Plot diagnostics for each fitted model
for (model_name in names(models)) {
  cat("Diagnostics for model:", model_name, "\n")
  plot_model_diagnostics(plot_data, models[[model_name]])
}
```

# 5. Output BEF
Export figure for main text and figure for supplementary materials

Select which BEF pairs to plot
Main figure: (1) FRic vs logC, (2) FDis vs logC, (3) FDiv vs ratioCN

## Fig 7
```{r}
# Functions for adding the curves in the plot
addFitCurve<- function(p, row, line_color) {
   coefficients <- sapply(strsplit(row$coefficients, ";")[[1]], function(x) {
    as.numeric(strsplit(x, "=")[[1]][2])
  })
  model_type <- row$best_model
  limits <- c(row$minX, row$maxX)
  
  if (model_type == "linear") {
    p <- p + stat_function(fun = function(x) coefficients[1] + coefficients[2] * x, 
                           xlim = limits,
                           size = 2, alpha = 0.8, color = line_color)
  } else if (model_type == "logarithmic") {
    p <- p + stat_function(fun = function(x) coefficients[1] + coefficients[2] * log(x),
                           xlim = limits,
                           size = 2, alpha = 0.8,  color = line_color)
  } else if (model_type == "power") {
    p <- p + stat_function(fun = function(x) coefficients[1] * x^coefficients[2], 
                           xlim = limits,
                           size = 2, alpha = 0.8,  color = line_color)
  } else if (model_type == "saturating") {
    p <- p + stat_function(fun = function(x) coefficients[1] * x / (coefficients[2] + x),
                           xlim = limits,
                           size = 2, alpha = 0.8,  color = line_color)
  }
  return(p)
}

# Function for plotting the data distribution and the fitted curves
plot_curves <- function(data, results.global, results.bioregion,
                        Div.val, CTV.val,
                        Div.Text = "",
                        CTV.Text = "") {
  # Subset data for this subplot
  plot_data <- data %>%
    filter(Diversity == Div.val, CTV == CTV.val)
  
  # Base plot
  p <- ggplot(plot_data, aes(x = Index, y = Value, color = Bioregion)) +
    geom_point(size = 1, alpha = 0.1) +
    scale_color_manual(values = clrs.bioregion) +  
    theme_bw() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
          legend.position = "none") +
    xlab(Div.Text) + ylab(CTV.Text)
  

  # Bioregion-specific curves
  row <- results.bioregion %>%
    filter(Diversity == Div.val, CTV == CTV.val)

  for (i in 1:nrow(row)) {
    p <- addFitCurve(p, row[i,], clrs.bioregion[i])
  }

  
  # Global curve
  row <- results.global %>% 
    filter(Diversity == Div.val, CTV == CTV.val)
  
  p <- addFitCurve(p, row, "black")
  
  
  return(p)
}

# Main figure
p1 <- plot_curves(data = BEF.2.mod, results.global, results.bioregion,
                  Div.val = "FRic", CTV.val = "logRespiration",
                  Div.Text = "Functional Richness",
                  CTV.Text = "Total Respiration Rate")


p2 <- plot_curves(data = BEF.2.mod, results.global, results.bioregion,
                  Div.val = "FDis", CTV.val = "logRespiration",
                  Div.Text = "Functional Dispersion",
                  CTV.Text = "Total Respiration Rate")


p3 <- plot_curves(data = BEF.2.mod, results.global, results.bioregion,
                  Div.val = "FDiv", CTV.val = "ratioCN",
                  Div.Text = "Functional Divergence",
                  CTV.Text = "CWM C:N")

cowplot::plot_grid(p1,p2,p3, nrow = 1, labels = "auto")

ggsave(filename = here("figures/fig_7_BEF.png"),
       width = 8, height= 7, units = "in", dpi = 600, bg = "white")

```


## Table S4 S5 Report selected curve fitting results in a table
Note that string parsing removes the negative signs for the b, so make sure to annotate them back.
```{r}
res.tab <- results.global %>% 
  mutate(CTV = factor(CTV, levels = c("logRespiration",
                                      "logExcretionN",
                                      "ratioCN"))) %>% 
  # filter(r2 >= 0.1) %>% 
  arrange(CTV, Diversity) %>% 
  # Isolate the values of a and b for the model
# note that for the linear model a = intercept, b = slope
  mutate(a = as.numeric(unlist(regmatches(coefficients, gregexpr("\\d+\\.\\d+", coefficients))))[1]) %>% 
  mutate(b = as.numeric(unlist(regmatches(coefficients, gregexpr("\\d+\\.\\d+", coefficients))))[2]) %>% 
  select(CTV, Diversity, best_model, a, b,
         aic, stderr, r2, coefficients)

# write.csv(res.tab, 
#           file = here("tables/curve_fit_global_results.csv"),
#           row.names = FALSE)


# Only report the pairs in the figures
res.tab.reg <- results.bioregion %>% 
  mutate(CTV = factor(CTV, levels = c("logRespiration",
                                      "logExcretionN",
                                      "ratioCN"))) %>% 
  # Select which bioregions to present
  filter((Diversity == "FRic" & CTV == "logRespiration") |
           (Diversity == "FDis" & CTV == "logRespiration") |
           (Diversity == "FDiv" & CTV == "ratioCN")) %>% 
  
  
  arrange(CTV, Diversity) %>% 
  group_by(Bioregion,CTV, Diversity) %>% 
  # Isolate the values of a and b for the model
# note that for the linear model a = intercept, b = slope
  mutate(a = as.numeric(unlist(regmatches(coefficients, gregexpr("\\d+\\.\\d+", coefficients))))[1]) %>% 
  mutate(b = as.numeric(unlist(regmatches(coefficients, gregexpr("\\d+\\.\\d+", coefficients))))[2]) %>% 
  select(CTV, Diversity, Bioregion, best_model, a, b,
         aic, stderr, r2, coefficients) %>% 
  ungroup()

# write.csv(res.tab.reg, 
#           file = here("tables/curve_fit_bioregion_results.csv"),
#           row.names = FALSE)
```
