---
title: "Functional group analysis"
author: "Patrick Pata"
date: "2023-10-24"
output: html_document
---

This file identifies the functional groups based on a cluster analysis of the gawdis dissimilarity between 8 traits. Figures exploring how the functional groups are projected in the trait space and the relationship of the other traits not used in the gawdis analysis to the trait space are generated. 

Additionally: 
- The significant differences in the means and ranges of trait values are compared between functional groups. 
- The proportion of functional groups to each of the association groups in Pata et al. (2022) are visualized.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load libraries and data
```{r}
library(tidyverse)
library(here)
library(cowplot)
library(scales)
set_here(path = "..")
`%notin%` <- Negate(`%in%`)

theme_set(theme_bw())

regions <- c("Offshore","DeepShelf","Nearshore","DeepFjord")
clrs.bioregion <- c("#004DFB","#FB20CA","#FFA500","#20FF00")

load(here("data/functional_grouping_results_20240617.RData"))

# Load a dataframe of merging the dendrogram at different levels
fg.merge <- read.csv(here("data/fg_cluster_merge.csv")) %>%
  rename(fungrp = fungrp10)
```

# 2. Create table of trait value averages between functional groups
```{r}
# Create table of species and numeric traits by functional grouo
sp.fg.num <- od.revised %>%
  select(Species, fungrp) %>%
  left_join(traits.num, 
            by = "Species") %>% 
  mutate(fungrp = paste0("FG",fungrp)) %>% 
  mutate(fungrp = factor(fungrp, levels = c("FG1","FG2","FG3","FG4","FG5",
                                            "FG6","FG7","FG8","FG9","FG10"))) %>% 
  arrange(fungrp)

# Longer dataframe
fungrp.num <- sp.fg.num %>% 
  pivot_longer(cols = -c(Species, fungrp), 
               names_to = "traitName", values_to = "traitValue") 


# Calculate mean, sd, median, and range for numeric traits
fg.traits.sum <- sp.fg.num %>% 
  pivot_longer(cols = -c(Species, fungrp),
               names_to = "traitName",
               values_to = "traitValue") %>% 
  group_by(fungrp, traitName) %>% 
  summarise(traitMean = mean(traitValue),
            traitSD = sd(traitValue),
            traitMedian = median(traitValue),
            traitIQR = IQR(traitValue),
            traitMin = min(traitValue),
            traitMax = max(traitValue),
            traitN = n(),
            .groups = "drop") %>% 
  arrange(traitName, fungrp)

# Export table of functional group by summary values, sectioned by traits. This will be annotated in excel.
write.csv(fg.traits.sum,
          file = here("tables/fg_traits_to_annotate.csv"), row.names = FALSE)

```

# 3. Calculate significant difference in trait values between functional groups
Note that the Dunns test will print out long outputs
```{r, eval=FALSE}
# Compare trait distributions between functional groups
library(broom)

# KW test comparing functional groups for each trait
fg.kw <- fungrp.num %>% 
  select(traitValue, traitName, fungrp) %>% 
  group_by(traitName) %>%
  do(tidy(kruskal.test(x=.$traitValue, g =.$fungrp))) %>% 
  mutate(fg.cluster = "FG10")


# Pairs of functional groups with significant difference based on Dunn test
fg.dun <- fungrp.num %>% 
  select(traitValue, traitName, fungrp) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.0005) %>% 
  mutate(fg.cluster = "FG10")

# Load a dataframe of merging the dendrogram at different levels
fg.merge <- read.csv(here("data/fg_cluster_merge.csv")) %>% 
  rename(fungrp = fungrp10)

# Group functional groups into merged groups
fungrp.num.2 <- fungrp.num %>% 
  left_join(fg.merge, by = "fungrp")

# It is only at the 2- and 3-cluster solutions when some traits return no significant difference between clusters
fg.kw.sub.3 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp3) %>% 
  group_by(traitName) %>%
  do(tidy(kruskal.test(x=.$traitValue, g =.$fungrp3))) %>% 
  filter(p.value < 0.05) %>% 
  arrange(-p.value) %>% 
  mutate(fg.cluster = "FG3")

# For 2-cluster solution, no need for Dunn test
fg.kw.sub.2 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp2) %>% 
  group_by(traitName) %>%
  do(tidy(kruskal.test(x=.$traitValue, g =.$fungrp2))) %>% 
  filter(p.value < 0.05) %>% 
  arrange(-p.value) %>% 
  mutate(fg.cluster = "FG2")



# Go through Dunn test for each level of clustering and rerun these tests
# Note that p.adjusted < 0.05 is retained here for reference but significant difference is really p < alpha/2
fg.dun.sub.3 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp3) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp3, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.05) %>% 
  arrange(-P.adjusted) %>% 
  mutate(fg.cluster = "FG3")

fg.dun.sub.4 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp4) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp4, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.05) %>% 
  arrange(-P.adjusted) %>% 
  mutate(fg.cluster = "FG4")

fg.dun.sub.5 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp5) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp5, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.05) %>% 
  arrange(-P.adjusted) %>% 
  mutate(fg.cluster = "FG5")

fg.dun.sub.6 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp6) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp6, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.05) %>% 
  arrange(-P.adjusted) %>% 
  mutate(fg.cluster = "FG6")

fg.dun.sub.7 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp7) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp7, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.05) %>% 
  arrange(-P.adjusted) %>% 
  mutate(fg.cluster = "FG7")

fg.dun.sub.8 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp8) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp8, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.05) %>% 
  arrange(-P.adjusted) %>% 
  mutate(fg.cluster = "FG8")

fg.dun.sub.9 <- fungrp.num.2 %>% 
  select(traitValue, traitName, fungrp9) %>% 
  group_by(traitName) %>%
  do(as.data.frame(dunn.test::dunn.test(x=.$traitValue, g =.$fungrp9, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted < 0.05) %>% 
  arrange(-P.adjusted) %>% 
  mutate(fg.cluster = "FG9")

fg.comp.kw.res <- bind_rows(
  fg.kw, fg.kw.sub.2, fg.kw.sub.3
)

fg.comp.dunn.res <- bind_rows(
  fg.dun, fg.dun.sub.9, fg.dun.sub.8, fg.dun.sub.7,
  fg.dun.sub.6, fg.dun.sub.5, fg.dun.sub.4, fg.dun.sub.3
)

save(fg.comp.kw.res, fg.comp.dunn.res,
     file = here("data/fg_comparison_kwdunn_results.RData"))

```

# 4. Plot results of KW and Dunn tests
```{r, warning=FALSE, fig.width=6}
# fg.comp.kw.res contains the overall fg KW and the 2- and 3-cluster solutions which have some no significant differences
# fg.com.dunn.res contains the Dunn test results with p-values < 0.05, although the significantly different pairs are those with p < alpha/2 (which is 0.0005).
load(here("data/fg_comparison_kwdunn_results.RData"))

# Select the main 10-cluster grouping
fg.comp.kw <- fg.comp.kw.res %>% 
  filter(fg.cluster == "FG10")

fg.comp.dunn <- fg.comp.dunn.res %>% 
  ungroup() %>% 
  filter(fg.cluster == "FG10") %>% 
  filter(P.adjusted < 0.0005) %>% 
  select(-fg.cluster) %>% 
  separate_wider_delim(cols = comparisons, delim = " - ",
                       names = c("FG.A","FG.B")) %>% 
  mutate(FG.A = factor(FG.A, levels = c("FG1","FG2","FG3","FG4","FG5",
                                        "FG6","FG7","FG8","FG9","FG10"))) %>% 
  mutate(FG.B = factor(FG.B, levels = c("FG1","FG2","FG3","FG4","FG5",
                                        "FG6","FG7","FG8","FG9","FG10")))
# Flip the instances when FG.A=10 to make all points all in one side of the triangle
fg.comp.dunn.1 <- fg.comp.dunn %>% 
  filter(FG.A != "FG10")
fg.comp.dunn.2 <- fg.comp.dunn %>% 
  filter(FG.A == "FG10") %>% 
  rename(FG.A = "FG.B", FG.B = "FG.A")
fg.comp.dunn <- bind_rows(fg.comp.dunn.1, fg.comp.dunn.2)
rm(fg.comp.dunn.1, fg.comp.dunn.2)

# Plot matrix of significant differences between functional group per trait
baseplot <- data.frame(FG.A = c("FG1","FG2","FG3","FG4","FG5",
                                "FG6","FG7","FG8","FG9","FG10")) %>%
  mutate(FG.A = factor(FG.A, levels = c("FG1","FG2","FG3","FG4","FG5",
                                        "FG6","FG7","FG8","FG9","FG10"))) %>% 
  mutate(FG.B = FG.A)

# Create a list of ggplots to selectively transform axes
traitplots <- vector("list",2)
count <- 1
tname_label = c("Water content","Body length","Dry weight","Carbon weight",
                "Ws. respiration rate","Ws. clearance rate",
                "Ws. ammonia excretion rate","Ws. phosphate excretion rate",
                "Ind. respiration rate","Ind. clearance rate",
                "Ind. ammonia excretion rate","Ind. phosphate excretion rate",
                "Nitrogen weight","Carbon content",
                "Nitrogen content","Phosphorus content",
                "Phosphorus weight","C:N","N:P","C:P")

for (tname in c("waterPWW","bodyLength","dryWeight","carbonWeight",
                "respirationRate_WSC", "clearanceRate_WSC",
                "excretionRateN_WSDW","excretionRateP_WSDW",
                "respirationRate","clearanceRate",
                "excretionRateN","excretionRateP",
                "nitrogenWeight","carbonPDW","nitrogenPDW","phosphorusPDW",
                "phosphorusWeight","ratioCN","ratioNP","ratioCP")) {
  traitplots[[count]] <- ggplot(baseplot, aes(x = FG.A, y = FG.B)) +
    geom_point(size = 0, shape = 3, color = "gray90") + 
    geom_point(data = fg.comp.dunn %>% 
                 filter(traitName == tname), 
               size = 3) +
    theme(axis.text.x = element_text(size = 12, face = "bold",
                                     color = fungrp.colors,
                                     angle = 45, vjust = 1, hjust=1),
          axis.text.y = element_text(size = 12, face = "bold",
                                     color = rev(fungrp.colors)),
          axis.title = element_blank(),
          plot.title = element_text(size = 10, face = "bold")) +
    scale_y_discrete(limits = rev) +
    ggtitle(tname_label[count]) 
  count <- count + 1
}

p.traits <- plot_grid(plotlist=traitplots, ncol = 4, labels = "auto")

# The number of traits that are significantly different between functional groups
fg.comp.dunn.sum <- fg.comp.dunn %>% 
  group_by(FG.A, FG.B) %>% 
  summarise(ntraits = n(),
            .groups = "drop")

p.overall <- ggplot(baseplot, aes(x = FG.A, y = FG.B)) +
  geom_point(size = 0, shape = 3, color = "gray90") + 
  geom_point(data = fg.comp.dunn.sum, 
             aes(size = ntraits)) +
  guides(size = guide_legend(title = 
                               "Number of\nsignificantly\ndifferent traits")) +
  theme(legend.position = "right",
        axis.text.x = element_text(size = 12, face = "bold",
                                   color = fungrp.colors),
        axis.text.y = element_text(size = 12, face = "bold",
                                   color = rev(fungrp.colors)),
        axis.title = element_blank(),
        plot.margin = margin(0.2,4,0.2,5, "cm")) +
    scale_y_discrete(limits = rev) 

plot_grid(p.traits, p.overall,
          ncol = 1, rel_heights = c(5,1.3),
          labels = c("","u"))

ggsave(file = here("figures/fig_s7__fungrps_sigdiff_dunn.png"),
       height = 16, width = 12, dpi = 300)

```


# 5. Comparing functional groups with association groups
```{r, eval=FALSE}
# Load list of association groups
#    Assoc group dataframe is sp.table from Pata et al. (2022; PinO)
load(here("data/Assoc_group_kendallresults_11022021.RData"))

# Match with functional groups
assoc.fg <- sp.table %>% 
  select(Species, AssocGroup, AssocGroup.sub) %>% 
  left_join(select(pcoa.vec, Species, fungrp, majorgroup),
            by = "Species") %>% 
  mutate(fungrp = if_else(is.na(fungrp), "no group", fungrp))

# view(assoc.fg %>% 
#        filter(is.na(fungrp)))
# assoc.fg.tab <- table(assoc.fg$AssocGroup, assoc.fg$fungrp)

assoc.fg.tab <- assoc.fg %>% 
  select(AssocGroup, fungrp) %>% 
  group_by(AssocGroup, fungrp) %>% 
  mutate(nspecies = n()) %>% 
  ungroup() %>% 
  distinct() %>% 
  mutate(fungrp = factor(fungrp, levels = c("FG1","FG2","FG3","FG4","FG5",
                                            "FG6","FG7","FG8","FG9","FG10",
                                            "no group"))) 

assoc.fg.tab.totals <- assoc.fg.tab %>% 
  group_by(AssocGroup) %>% 
  mutate(totalNG = sum(nspecies)) %>% 
  ungroup() %>% 
  filter(fungrp != "no group") %>% 
  group_by(AssocGroup) %>% 
  mutate(totalFG = sum(nspecies)) %>% 
  ungroup() %>% 
  distinct(AssocGroup, totalNG, totalFG)

assoc.fg.tab <- assoc.fg.tab %>% 
  left_join(assoc.fg.tab.totals) %>% 
  mutate(AGlab1 = paste0(AssocGroup, '\nN = ', totalNG),
         AGlab2 = paste0(AssocGroup, '\nN = ', totalFG))
  
# Plot
ggplot(assoc.fg.tab,
             aes(x = nspecies, y = AGlab1, fill = fungrp)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_x_reverse(position = "bottom", expand = c(0, 0),
                  labels = c("0.00" = "1","0.25" = "0.75",
                             "0.50" = "0.50", "0.75" = "0.25", "1.00" = "0")) +
  ylab("Association Group") + xlab("Proportion") +
  scale_y_discrete(limits = rev) +
  scale_fill_manual("Functional Group",
    values = c(fungrp.colors,"cornsilk3")) +
  guides(fill=guide_legend(nrow=2,byrow=TRUE)) +
  theme(legend.position = "top", legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(0,0,0,-10))

# ggsave(filename = here("figures/extra_assoc_fun_group_match.png"),
#     width = 8, height= 4, units = "in", dpi = 600, bg = "white")


# Plot version 2 without "no group" species
ggplot(assoc.fg.tab %>% filter(fungrp != "no group"),
             aes(x = nspecies, y = AGlab2, fill = fungrp)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_x_reverse(position = "bottom", expand = c(0, 0),
                  labels = c("0.00" = "1","0.25" = "0.75",
                             "0.50" = "0.50", "0.75" = "0.25", "1.00" = "0")) +
  ylab("Association Group") + xlab("Proportion") +
  scale_y_discrete(limits = rev) +
  scale_fill_manual("Functional Group",
    values = c(fungrp.colors,"cornsilk3")) +
  guides(fill=guide_legend(nrow=2,byrow=TRUE)) +
  theme(legend.position = "top", legend.justification = c(0,1),
        legend.margin=margin(0,0,0,0), legend.box.margin=margin(0,0,0,-10))

# ggsave(filename = here("figures/extra_assoc_fun_group_match_v2.png"),
#     width = 8, height= 4, units = "in", dpi = 600, bg = "white")
```

