---
title: "Bioregion compositon by functional group"
author: "Patrick Pata"
date: "2023-11-02"
output: html_document
---

This file calculates the functional group composition of the bioregions. The relative abundance and biomass of functional groups in trait space is also visualized. Bioregions are compared according to the absolute and relative abundance and biomass of functional groups overall, by season, and interannually.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load libraries and data
```{r}
library(tidyverse)
library(here)
library(data.table) # for calculating convex hulls
library(scales)
library(ggtext)
library(ggh4x) # for strip colors
library(cowplot)
library(broom)
set_here(path = "..")
`%notin%` <- Negate(`%in%`)

theme_set(theme_bw())

# The original community data
load(here("data/zoop_data_regionalized_annotated_20231018.RData"))
regions <- c("Offshore","DeepShelf","Nearshore","DeepFjord")

# Functional group data + traits
load(here("data/functional_grouping_results_20240617.RData"))

geomean <- function(x) {  exp(mean(log(x), na.rm = TRUE))  }
```



# 2. Bioregion relative composition
Calculate the total abundance and biomass of each functional group between bioregions
```{r, fig.width=5}
strip_fg <- ggh4x::strip_themed(background_x = ggh4x::elem_list_rect(
  fill = gplots::col2hex(fungrp.colors)))

# Filter the community data to only include the 163 species with trait information
zoop.bc.species.fun <- zoop.bc %>% 
  # Add fuctional group information
  left_join(select(od.revised, Species, fungrp),
            by = "Species") %>% 
  filter(!is.na(fungrp)) %>% 
  # Convert density to by volume
  left_join(select(zoopmeta, Key, Tow.Depth), by = "Key") %>% 
  mutate(Abundance.m3 = Abundance.m2 / Tow.Depth,
         Biomass.mg.m3 = Biomass.mg.m2 / Tow.Depth) %>% 
  select(Key, Bioregion, Species, 
         majorgroup, fungrp, 
         Tow.Depth,
         Abundance.m2, Biomass.mg.m2,
         Abundance.m3, Biomass.mg.m3)

zoop.bc.fun <- zoop.bc.species.fun %>% 
  group_by(Key, Tow.Depth, Bioregion, fungrp) %>% 
  summarise(Abundance.m2 = sum(Abundance.m2), Abundance.m3 = sum(Abundance.m3),
         Biomass.mg.m2 = sum(Biomass.mg.m2), Biomass.mg.m3 = sum(Biomass.mg.m3), 
         .groups = "drop") %>% 
  # Calculate relative abundance and biomass
  group_by(Key, Bioregion) %>% 
  mutate(Total.Biomass = sum(Biomass.mg.m2),
         Total.Abundance = sum(Abundance.m2)) %>% 
  ungroup() %>% 
  mutate(Percent.Biomass = Biomass.mg.m2/Total.Biomass*100,
         Percent.Abundance = Abundance.m2/Total.Abundance*100) %>% 
  select(-c(Total.Biomass, Total.Abundance)) 
  
# save(zoop.bc.species.fun, zoop.bc.fun,
#      file = here("data/key_fg_abundance_biomass_20240626.RData"))
```



# 3. Trait space "heatmaps"
Color by functional group, size according to abundance or  biomass


## Prepare the dataframes
```{r}
# Extract the pcoa table from pcoa.vec
species.pcoa <- pcoa.vec[,c(1:6)] %>% 
  column_to_rownames("Species")


# Calculate average species abundance and biomass for each bioregion
bioregion.biom <- zoop.bc.species.fun %>% 
    select(Key, Bioregion, Species, Biomass.mg.m2) %>% 
  group_by(Bioregion) %>% 
  mutate(nsamples = length(unique(Key))) %>% 
  ungroup() %>% 
  pivot_wider(names_from = Species, values_from = Biomass.mg.m2,
              values_fill = 0) %>% 
  pivot_longer(cols = -c(Key, Bioregion, nsamples),
               names_to = "Species", values_to = "Biomass.mg.m2") %>% 
  mutate(Presence = if_else(Biomass.mg.m2 > 0, 1, 0)) %>% 
  group_by(Bioregion, Species, nsamples) %>% 
  summarise(Biomass = mean(Biomass.mg.m2),
            Presence = sum(Presence) /nsamples*100, .groups = "drop") %>% 
  distinct()

bioregion.abun <- zoop.bc.species.fun %>% 
    select(Key, Bioregion, Species, Abundance.m2) %>% 
  group_by(Bioregion) %>% 
  mutate(nsamples = length(unique(Key))) %>% 
  ungroup() %>% 
  pivot_wider(names_from = Species, values_from = Abundance.m2,
              values_fill = 0) %>% 
  pivot_longer(cols = -c(Key, Bioregion, nsamples),
               names_to = "Species", values_to = "Abundance.m2") %>% 
  mutate(Presence = if_else(Abundance.m2 > 0, 1, 0)) %>% 
  group_by(Bioregion, Species, nsamples) %>% 
  summarise(Abundance = mean(Abundance.m2),
            Presence = sum(Presence) /nsamples*100, .groups = "drop") %>% 
  distinct()

# Add this to pcoa.vec - This will exclude taxa not included in the functional group analysis
sp.regions <- pcoa.vec %>% 
  left_join(select(bioregion.biom, Species, Bioregion, Biomass, Presence), 
            by = c("Species")) %>% 
  left_join(select(bioregion.abun, Species, Bioregion, Abundance), 
            by = c("Species","Bioregion")) %>% 
  # Modify bioregion names to include spaces
  mutate(Bioregion = if_else(Bioregion == "DeepShelf", 
                             "Deep Shelf", Bioregion),
         Bioregion = if_else(Bioregion == "DeepFjord", 
                             "Deep Fjord", Bioregion)) %>% 
  mutate(Bioregion = factor(Bioregion,
                            levels = c("Offshore","Deep Shelf",
                                       "Nearshore","Deep Fjord"))) %>% 

  mutate(fungrp = factor(fungrp, levels = c("FG1","FG2","FG3","FG4","FG5",
                                            "FG6","FG7","FG8","FG9","FG10"))) %>% 
  arrange(-Biomass)
```


## PCoA 1 & 2: Abundance + Biomass
```{r}
axestoplot <- c(1,2)
# Create the global and regional hulls
clustPoints <- as.data.frame(species.pcoa[,axestoplot]) %>%
  mutate(group = "all") 
dt <- data.table(xdata=clustPoints[,1], ydata=clustPoints[,2], 
                 level=clustPoints[,3], key="level")
hulls.global <- dt[dt[, .I[chull(xdata, ydata)], by = level]$V1]


hplot.abund <- ggplot(data = sp.regions %>% 
         filter(Presence >= 0), aes(x = PC1, y = PC2)) +
    geom_polygon(data = hulls.global, aes(x=xdata,y=ydata,fill=level),
                 color = "white", fill = "white", show.legend = FALSE) +

    # Distribution of species
    geom_point(data = sp.regions %>% 
         filter(Presence >= 0),
               aes(color = fungrp, size = Abundance),
               alpha = 0.5) +
    scale_color_manual(values = fungrp.colors,
                       name = "Functional Groups") +
    
    scale_size_area(max_size = 20, breaks=c(10,100,1000,10000,20000),
                    name = "Abundance\n(indiv m-2)") +
  
    theme(aspect.ratio = 1, legend.position="right", 
          legend.box = "veritcal",
          panel.background = element_rect(fill = "grey80"),
          strip.text = element_text(size = 10)) +
  guides(color = guide_legend(title.position = "top", 
                              title.hjust = 0.5, ncol = 2,
                              override.aes = list(size = 3),
                              order = 2),
         size = guide_legend(title.position = "top", 
                              title.hjust = 0.5,
                             order = 1)) +
  xlab("PCoA 1") + ylab("PCoA 2") +
  facet_wrap(~Bioregion, ncol = 2)

hplot.abund
ggsave(filename = here("figures/fig_4_traits_space_abundance_PC12.png"),
    width = 7, height= 6, units = "in", dpi = 600, bg = "white")

hplot.biom <- ggplot(data = sp.regions %>% 
         filter(Presence >= 0), aes(x = PC1, y = PC2)) +
    geom_polygon(data = hulls.global, aes(x=xdata,y=ydata,fill=level),
                 color = "white", fill = "white", show.legend = FALSE) +
  
    # Distribution of species
    geom_point(data = sp.regions %>% 
         filter(Presence >= 0),
               aes(color = fungrp, size = Biomass),
               alpha = 0.5) +
    scale_color_manual(values = fungrp.colors,
                       name = "Functional Groups") +
    
    scale_size_area(max_size = 23, breaks=c(10,100,1000,2000,5000),
                    name = "Biomass\n(mg m-2)") +
  
    theme(aspect.ratio = 1, legend.position="right", 
          legend.box = "veritcal",
          panel.background = element_rect(fill = "grey80"),
          strip.text = element_text(size = 10)) +
  guides(color = guide_legend(title.position = "top", 
                              title.hjust = 0.5, ncol = 2,
                              override.aes = list(size = 3),
                              order = 2),
         size = guide_legend(title.position = "top", 
                              title.hjust = 0.5,
                             order = 1)) +
  xlab("PCoA 1") + ylab("PCoA 2") +
  facet_wrap(~Bioregion, ncol = 2)

hplot.biom
ggsave(filename = here("figures/fig_5_traits_space_biomass_PC12.png"),
    width = 7, height= 6, units = "in", dpi = 600, bg = "white")

```

# 4.Table of bioregion comparisons
```{r}
zoop.bc.fun.zeros <- zoop.bc.fun %>% 
  # Create a blank table of Key and functional groups
  distinct(Key, Bioregion) %>% 
  left_join(distinct(zoop.bc.fun, Bioregion, fungrp),
            by = "Bioregion") %>% 
  # Add the non-zero instances
  left_join(select(zoop.bc.fun,
                   Key, Bioregion, fungrp,
                   Abundance.m2, Biomass.mg.m2,
                   Percent.Abundance, Percent.Biomass),
            join_by(Key, Bioregion, fungrp)) %>% 
  # Convert NAs to zeros
  mutate(Abundance.m2 = if_else(is.na(Abundance.m2), 0, Abundance.m2),
         Biomass.mg.m2 = if_else(is.na(Biomass.mg.m2), 0 , Biomass.mg.m2),
         Percent.Abundance = if_else(is.na(Percent.Abundance), 
                                     0, Percent.Abundance),
         Percent.Biomass = if_else(is.na(Percent.Biomass),
                                   0, Percent.Biomass))

zoop.bc.sum <- zoop.bc.fun.zeros %>% 
  select(-Key) %>% 
  pivot_longer(cols = -c(Bioregion, fungrp),
               names_to = "Metric", values_to = "Value") %>% 
  group_by(Bioregion, fungrp, Metric) %>% 
  summarise(ValMean = mean(Value),
            ValSD = sd(Value), 
            ValMedian = median(Value),
            ValIQR = IQR(Value),
            ValMin = min(Value),
            ValMax = max(Value),
            ValN = n(),
            .groups = "drop") %>% 
  arrange(Metric, fungrp)

# Per bioregion and functional group
zoop.bc.sum.table <- zoop.bc.sum %>% 
  filter(Metric %notin% c("Abundance.m3","Biomass.mg.m3")) %>% 
  select(Bioregion, fungrp, Metric, ValMean, ValSD) %>% 
  mutate(ValText = if_else(Metric %in% c("Percent.Abundance",
                                          "Percent.Biomass"), 
           paste(round(ValMean,2),"\u00b1",round(ValSD,1)),
           paste(round(ValMean,0),"\u00b1",round(ValSD,0)))) %>% 
  select(-c(ValMean, ValSD)) %>% 
  pivot_wider(names_from = Bioregion,
              values_from = ValText) 

# Total for all
zoop.bc.total <- zoop.bc.fun.zeros %>% 
  group_by(Key, Bioregion) %>% 
  summarise(Total.Abundance = sum(Abundance.m2),
            Total.Biomass = sum(Biomass.mg.m2),
            .groups = "drop") %>% 
  group_by(Bioregion, ) %>% 
  summarise(Abun.Mean = mean(Total.Abundance),
            Abun.SD = sd(Total.Abundance), 
            Biom.Mean = mean(Total.Biomass),
            Biom.SD = sd(Total.Biomass), 
            N = n(),
            .groups = "drop") %>% 
  mutate(Abundance.Total = paste(round(Abun.Mean,0),"\u00b1",round(Abun.SD,0)),
         Biomass.Total = paste(round(Biom.Mean,0),"\u00b1",round(Biom.SD,0))) %>% 
  select(Bioregion, Abundance.Total, Biomass.Total) %>% 
  pivot_longer(cols = c("Abundance.Total","Biomass.Total"),
               names_to = "Metric") %>% 
  pivot_wider(names_from = Bioregion, values_from = value)
  
zoop.bc.sum.table <- bind_rows(zoop.bc.total,zoop.bc.sum.table) %>% 
  relocate(Metric, fungrp)


# # Export table for annotation in excel.
# write.csv(zoop.bc.sum.table,
#           file = here("tables/bioregion_fg_means_to_annotate_20240704.csv"), row.names = FALSE)

```


## Relative abundance and biomass
```{r}
# As violin plots
g1 <- ggplot(zoop.bc.fun.zeros, aes(x = Bioregion, y = Percent.Abundance, 
                              fill = Bioregion)) +
  geom_violin(aes(color = Bioregion), alpha = 0.6) + 
  geom_boxplot(width = 0.2, color = "grey25", outlier.shape = NA) +
  # stat_summary(fun = "mean", geom = "crossbar", width = 0.5,
  #              colour = "black") +
  xlab("") +
  ylab("Percent Abundance") +
  theme(legend.position = "none",  axis.title.y = element_markdown(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  scale_fill_manual(values = clrs.bioregion) +
  scale_color_manual(values = clrs.bioregion) +
  # scale_y_continuous(trans = "log10", 
  #                    breaks = c(0.001,0.01,0.1,1,10,100,1000,10000,100000,1000000),
  #                    labels = c("10^-3","10^-2","10^-1","10^0","10^1",
  #                               "10^2","10^3","10^4","10^5","10^6")) +
  facet_wrap(~fungrp,  nrow = 1) +
  ggh4x::facet_wrap2(~fungrp, nrow = 1, strip = strip_fg)

g2 <- ggplot(zoop.bc.fun.zeros, aes(x = Bioregion, y = Percent.Biomass, 
                              fill = Bioregion)) +
  geom_violin(aes(color = Bioregion), alpha = 0.6) + 
  geom_boxplot(width = 0.2, color = "grey25", outlier.shape = NA) +
  # stat_summary(fun = "mean", geom = "crossbar", width = 0.5,
  #              colour = "black") +
  xlab("") +
  ylab("Percent Biomass") +
  theme(legend.position = "none",  axis.title.y = element_markdown(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  scale_fill_manual(values = clrs.bioregion) +
  scale_color_manual(values = clrs.bioregion) +
  # scale_y_continuous(trans = "log10", 
  #                    breaks = c(0.001,0.01,0.1,1,10,100,1000,10000,100000,1000000),
  #                    labels = c("10^-3","10^-2","10^-1","10^0","10^1",
  #                               "10^2","10^3","10^4","10^5","10^6")) +
  facet_wrap(~fungrp,  nrow = 1) +
  ggh4x::facet_wrap2(~fungrp, nrow = 1, strip = strip_fg)

legend_b <- get_legend(g1 +    
                         guides(color = guide_legend(nrow = 1)) +
                         theme(legend.position = "bottom"))

plot_grid(plot_grid(g1, g2, labels = "auto", ncol = 1), 
          legend_b, nrow = 2, rel_heights = c(1,0.05))


# ggsave(filename = here("figures/fig_violin_percent_contribution_region.png"),
# width = 10, height= 5, units = "in", dpi = 600, bg = "white")



# As stacked bar plots
zbs.sum <- zoop.bc.sum %>% 
  filter(Metric %in% c("Percent.Abundance",
                       "Percent.Biomass")) %>% 
  mutate(Metric = if_else(Metric == "Percent.Abundance",
                 "% Abundance", "% Biomass")) %>% 
  mutate(fungrp = as.factor(fungrp))
  

ggplot(zbs.sum, aes(y = ValMean, x = Bioregion, 
                     fill = fungrp)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(values = fungrp.colors, 
                    name = "Functional\ngroup") +
  scale_y_continuous(expand = c(0, 0),
                  labels = c("0.00" = "0%","0.25" = "25%",
                             "0.50" = "50%", "0.75" = "75%", "1.00" = "100%")) +
  facet_wrap(~Metric) 


# ggsave(filename = here("figures/fig_bar_percent_contribution_region.png"),
# width = 7, height= 4, units = "in", dpi = 600, bg = "white")

```

## Plot as treemap for summary figure 9
```{r}
# install.packages("treemapify")
# library(treemapify)

ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Biomass") %>% 
         filter(Bioregion == "Offshore"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")

ggsave(filename = here("figures/fig9/treemap_biomass_offshore.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")


ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Biomass") %>% 
         filter(Bioregion == "DeepShelf"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")


ggsave(filename = here("figures/fig9/treemap_biomass_deepshelf.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")


ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Biomass") %>% 
         filter(Bioregion == "Nearshore"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")

ggsave(filename = here("figures/fig9/treemap_biomass_nearshore.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")

ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Biomass") %>% 
         filter(Bioregion == "DeepFjord"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")

ggsave(filename = here("figures/fig9/treemap_biomass_deepfjord.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")

ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Abundance") %>% 
         filter(Bioregion == "DeepFjord"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")

ggsave(filename = here("figures/fig9/treemap_abundance_deepfjord.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")


ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Abundance") %>% 
         filter(Bioregion == "Nearshore"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")

ggsave(filename = here("figures/fig9/treemap_abundance_nearshore.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")

ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Abundance") %>% 
         filter(Bioregion == "DeepShelf"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")

ggsave(filename = here("figures/fig9/treemap_abundance_deepshelf.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")


ggplot(zoop.bc.sum %>% 
         filter(Metric == "Percent.Abundance") %>% 
         filter(Bioregion == "Offshore"), 
       aes(area = ValMean, fill = as.factor(fungrp))) +
  treemapify::geom_treemap(color = "white") +
  treemapify::geom_treemap_text(aes(label = paste0("FG",fungrp)), 
                                colour = "white") +
  theme(legend.position = "none") +
  scale_fill_manual(values = fungrp.colors) +
  theme(legend.position = "none")

ggsave(filename = here("figures/fig9/treemap_abundance_offshore.png"),
width = 4, height= 4, units = "in", dpi = 600, bg = "white")
```


# 5. Significance tests
## Sig diff test Overall
```{r}
# Overall total
zoop.bc.total <- zoop.bc.fun.zeros %>% 
  group_by(Key, Bioregion) %>% 
  summarise(Total.Abundance = sum(Abundance.m2),
            Total.Biomass = sum(Biomass.mg.m2),
            .groups = "drop")
kruskal.test(Total.Abundance ~ Bioregion, data = zoop.bc.total)
kruskal.test(Total.Biomass ~ Bioregion, data = zoop.bc.total)

dunn.test::dunn.test(x=zoop.bc.total$Total.Abundance, 
                     g =zoop.bc.total$Bioregion, 
                     method = "bh", alpha = 0.001)

dunn.test::dunn.test(x=zoop.bc.total$Total.Biomass, 
                     g =zoop.bc.total$Bioregion, 
                     method = "bh", alpha = 0.001)
```


## Sig diff test Total
```{r, eval=FALSE}
library(broom)

colnames(zoop.bc.fun.zeros)

# Pivot wider to apply tests to every columns
fun.biom <- zoop.bc.fun.zeros %>% 
  # mutate(Biomass.mg.m3 = log1p(Biomass.mg.m3)) %>% 
  select(Key, Bioregion, fungrp, Biomass.mg.m2) %>% 
  pivot_wider(names_from = fungrp, values_from = Biomass.mg.m2, 
              values_fill = NA) %>% 
  column_to_rownames("Key")
fun.abun <- zoop.bc.fun.zeros %>% 
  # mutate(Abundance.m3 = log1p(Abundance.m3)) %>% 
  select(Key, Bioregion, fungrp, Abundance.m2) %>% 
  pivot_wider(names_from = fungrp, values_from = Abundance.m2, 
              values_fill = NA) %>% 
  column_to_rownames("Key")

# presence of functional groups in each bioregion
fun.pres <- zoop.bc.fun.zeros %>% 
  select(Key, Bioregion, fungrp, Biomass.mg.m2) %>% 
  filter(!is.na(Biomass.mg.m2)) %>% 
  group_by(fungrp) %>% 
  count()

# KW test per functional group
fun.biom.kw <- fun.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(tidy(kruskal.test(x=.$value, g =.$Bioregion)))

fun.abun.kw <- fun.abun %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(tidy(kruskal.test(x=.$value, g =.$Bioregion)))


# post-hoc dunn test
fun.biom.dun <- fun.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$value, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted >= 0.0005)
fun.abun.dun <- fun.abun %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$value, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted >= 0.0005)

save(fun.biom.kw, fun.abun.kw,
     fun.biom.dun, fun.abun.dun,
     file = here("data/abundance_biomass_kw_dunn_results_m2_N3721_20240704.RData"))

```

## Sig diff test Percent
```{r, eval=FALSE}

# Pivot wider to apply tests to every columns
fun.biom.perc <- zoop.bc.fun.zeros %>% 
  select(Key, Bioregion, fungrp, Percent.Biomass) %>% 
  pivot_wider(names_from = fungrp, values_from = Percent.Biomass, 
              values_fill = NA) %>% 
  column_to_rownames("Key")

fun.abun.perc <- zoop.bc.fun.zeros %>% 
  select(Key, Bioregion, fungrp, Percent.Abundance) %>% 
  pivot_wider(names_from = fungrp, values_from = Percent.Abundance, 
              values_fill = NA) %>% 
  column_to_rownames("Key")


# KW test per functional group
fun.biom.perc.kw <- fun.biom.perc %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(tidy(kruskal.test(x=.$value, g =.$Bioregion)))

fun.abun.perc.kw <- fun.abun.perc %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(tidy(kruskal.test(x=.$value, g =.$Bioregion)))


# post-hoc dunn test
fun.biom.perc.dun <- fun.biom.perc %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$value, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted >= 0.0005)

fun.abun.perc.dun <- fun.abun.perc %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$value, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted >= 0.0005)

save(fun.biom.perc.kw, fun.abun.perc.kw,
     fun.biom.perc.dun, fun.abun.perc.dun,
     file = here("data/abundance_biomass_percent_kw_dunn_results_m2_N3721_20240704.RData"))

# Most common similarities between functional groups
dunn.sum <- bind_rows(fun.abun.dun, fun.biom.dun,
                      fun.abun.perc.dun, fun.biom.perc.dun) %>% 
  group_by(comparisons) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  arrange(-n)
```


