---
title: "Comparison of bioregions"
author: "Patrick Pata"
date: "2024-06-25"
output: html_document
---

This file calculates the community level trait totals and community weighted means for each sample. These can be recalculated with the species-level community data but the results of the calculations are provided in the community_metrics_*.RData file to allow further visualization and analysis.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# 1. Load libraries and data
```{r}
library(tidyverse)
library(here)
library(ggtext)
library(scales)
library(cowplot)
library(broom)

set_here(path = "..")
`%notin%` <- Negate(`%in%`)
theme_set(theme_bw())

# Functional group data + traits
load(here("data/functional_grouping_results_20240617.RData"))

# FG density, CWM, and community total data
load(here("data/community_metrics_3721_163species_20250217.RData"))

regions <- c("Offshore","DeepShelf","Nearshore","DeepFjord")
clrs.bioregion <- c("#004DFB","#FB20CA","#FFA500","#20FF00")


fungrp.colors <- c("turquoise3", "steelblue2","darkorange2",
                   "orchid","magenta4","lightcoral",
                   "gold3","darkgreen","darkolivegreen4",
                   "olivedrab3")
                   
standard_temp <- function(rate0, t0, t = 15, Q10 = 2.8) {
  10^(log10(rate0) + log10(Q10)*((t-t0)/10) )
}
```


# 2. Community level traits
## Calculate CWM and proportions of traits
```{r, eval=FALSE}
# Load data of samples with updated bioregions and functional groups
load(here("data/restricted_data/samples_species_fg_abundance_biomass_20250217.RData"))

# Organize the trait data
all.traits <- species.traits[,6:15] %>% 
  rownames_to_column("Species") %>% 
  left_join(traits.num, by = "Species") %>% 
  pivot_longer(cols = -Species, 
               names_to = "traitName",
               values_to = "traitValue")

zoop.163.density <- zoop.bc.species.fun %>% 
  filter(Species %in% all.traits$Species) %>%  
  group_by(Key, Tow.Depth, Bioregion, fungrp) %>% 
  summarise(Abundance.m2 = sum(Abundance.m2), Abundance.m3 = sum(Abundance.m3),
         Biomass.mg.m2 = sum(Biomass.mg.m2), Biomass.mg.m3 = sum(Biomass.mg.m3), 
         .groups = "drop")

zoop.163.cwm <- zoop.bc.species.fun %>% 
  filter(Species %in% all.traits$Species) %>% 
  left_join(all.traits, by = "Species") %>% 
  rename(Trait = traitName) %>% 
  # Calculate community weighted mean
  group_by(Key, Bioregion, Trait) %>% 
  summarise(CWM = weighted.mean(traitValue, Biomass.mg.m2),
            .groups = "drop")
```

## Calculate community trait totals
```{r, eval=FALSE}
# Load the complete community data for community totals
load(here("data/restricted_data/BC_3721_163_speciesstage_20241024.RData"))

all.traits.wide <- all.traits %>% 
  filter(traitName %in% c("dryWeight","carbonWeight",
                      "nitrogenWeight","phosphorusWeight",
                      "respirationRate","clearanceRate",
                      "excretionRateN","excretionRateP")) %>% 
  pivot_wider(names_from = traitName, values_from = traitValue)
  

# Actually calculate weights and individual rates from % and weight specific
zoop.spstg.ctotal <- zoop.fg.com %>% 
  mutate(dryWeight = `Biomass(mg/m3)` / `Abundance(#/m3)`) %>% 
  left_join(traits.num %>% 
              select(Species, waterPWW, carbonPDW, nitrogenPDW, phosphorusPDW,
                     respirationRate_WSC, clearanceRate_WSC, 
                     excretionRateN_WSDW, excretionRateP_WSDW,
                     ratioCN, ratioNP, ratioCP, bodyLength),
            join_by(Species == Species)) %>% 
  # This filter will effectively exclude all species-stage records that were not included in the taxonomic bioregionalization
  filter(!is.na(carbonPDW)) %>% 
  # Calculate C,N,P weights
  mutate(carbonWeight = dryWeight * carbonPDW/100,
         nitrogenWeight = dryWeight * nitrogenPDW/100,
         phosphorusWeight = dryWeight * phosphorusPDW/100) 
 
# If temp.mean is not present, use the bioregion averages
bioreg.temp <- zoop.spstg.ctotal %>% 
  filter(!is.na(temp.mean)) %>% 
  distinct(Bioregion, Key, temp.mean) %>% 
  group_by(Bioregion) %>% 
  summarise(RegionTemp = mean(temp.mean), 
            .groups = "drop") 

zoop.spstg.ctotal <- zoop.spstg.ctotal %>%
  left_join(bioreg.temp) %>% 
  mutate(temp.mean = if_else(is.na(temp.mean), RegionTemp, temp.mean)) %>% 
  # Adjust rates to the net integrated temperature
  mutate(respirationRate_WSC = standard_temp(respirationRate_WSC, t0 = 8.5,
                                             t = temp.mean, Q10 = 2.8),
         clearanceRate_WSC = standard_temp(clearanceRate_WSC, t0 = 8.5,
                                           t = temp.mean, Q10 = 2.8),
         excretionRateN_WSDW = standard_temp(excretionRateN_WSDW, t0 = 8.5,
                                             t = temp.mean, Q10 = 2.0),
         excretionRateP_WSDW = standard_temp(excretionRateP_WSDW, t0 = 8.5,
                                             t = temp.mean, Q10 = 1.55)) %>%
  # Calculate individual specific rates
  mutate(respirationRate = carbonWeight * respirationRate_WSC,
         clearanceRate = carbonWeight * clearanceRate_WSC, 
         excretionRateN = dryWeight * excretionRateN_WSDW,
         excretionRateP = dryWeight * excretionRateP_WSDW) 

# save(zoop.spstg.ctotal,
#   file = here("data/restricted_data/BC_3721_163_speciesstage_traits_20250217.RData"))

# Calculate the total community level traits and abundance + biomass for each sample
zoop.163.ctotal <- zoop.spstg.ctotal %>%
  mutate(Abundance.m2 = `Abundance(#/m3)` * Depth_Start) %>% 
  mutate(Biomass.mg.m2 = `Biomass(mg/m3)` * Depth_Start) %>% 
  
  # Calculate community totals
  group_by(Key, Bioregion, fungrp) %>%
  summarise(totalDryWeight = sum(dryWeight * Abundance.m2),
            totalCarbonWeight = sum(carbonWeight * Abundance.m2),
            totalNitrogenWeight = sum(nitrogenWeight * Abundance.m2),
            totalPhosphorusWeight = sum(phosphorusWeight * Abundance.m2),
            totalRespirationRate = sum(respirationRate * Abundance.m2),
            totalClearanceRate = sum(clearanceRate * Abundance.m2),
            totalExcretionRateN = sum(excretionRateN * Abundance.m2),
            totalExcretionRateP = sum(excretionRateP * Abundance.m2),
            Abundance = sum(Abundance.m2),
            Biomass = sum(Biomass.mg.m2),
            .groups = "drop") %>% 
 
  # Lengthen
  pivot_longer(cols = c(totalDryWeight:Biomass),
               names_to = "Trait", values_to = "CTotal") %>% 

  # Sum for each samples
  group_by(Key, Bioregion, Trait) %>% 
  summarise(CTotal = sum(CTotal), .groups = "drop")
 

# Community total traits by functional group 
zoop.163.ctotal.ave <- zoop.spstg.ctotal %>%
  mutate(Abundance.m2 = `Abundance(#/m3)` * Depth_Start) %>% 
  mutate(Biomass.mg.m2 = `Biomass(mg/m3)` * Depth_Start) %>% 
  
  # Calculate community totals
  group_by(Key, Bioregion, fungrp) %>%
  summarise(dryWeight = sum(dryWeight * Abundance.m2),
            carbonWeight = sum(carbonWeight * Abundance.m2),
            nitrogenWeight = sum(nitrogenWeight * Abundance.m2),
            phosphorusWeight = sum(phosphorusWeight * Abundance.m2),
            respirationRate = sum(respirationRate * Abundance.m2),
            clearanceRate = sum(clearanceRate * Abundance.m2),
            excretionRateN = sum(excretionRateN * Abundance.m2),
            excretionRateP = sum(excretionRateP * Abundance.m2),
            Abundance = sum(Abundance.m2),
            Biomass = sum(Biomass.mg.m2),
            .groups = "drop") %>% 
  
  # pad zeros to empty FGs (not really necessary... because the Key-Bioregion-FG pair with blanks will never show up (e.g., FG2))
  pivot_longer(cols = c(dryWeight:Biomass),
               names_to = "Metric", values_to = "Value") %>% 
  pivot_wider(names_from = Metric, values_from = Value,
              values_fill = 0) %>%
  pivot_longer(cols = c(dryWeight:Biomass),
               names_to = "Metric", values_to = "Value") %>% 
  
  # Average across all values (not staggered by year-month)
  group_by(Bioregion, Metric, fungrp) %>% 
  summarise(Value.mean = mean(Value),
            Value.sd = sd(Value),
            nsamples = n(), .groups = "drop") %>% 
  mutate(Value.se = Value.sd / sqrt(nsamples)) %>% 
  mutate(fungrp = factor(fungrp, levels = c("1","2","3","4","5","6",
                                            "7","8","9","10")))


# Export values
save(zoop.163.density, zoop.163.cwm, zoop.163.ctotal, zoop.163.ctotal.ave,
  file = here("data/community_metrics_3721_163species_20250217.RData"))
```



# 3. Fig4 Community total trait values and average contribution of the functional groups
```{r}
zoop.163.ctotal.plot <- zoop.163.ctotal.ave %>% 
  filter(Metric %notin% c("Biomass","nitrogenWeight", 
                          "phosphorusWeight", "dryWeight")) %>% 
  mutate(Metric = if_else(Metric == "carbonWeight",
                          "Carbon Weight", Metric),
         Metric = if_else(Metric == "clearanceRate",
                          "Clearance Rate", Metric),
         Metric = if_else(Metric == "respirationRate",
                          "Respiration Rate", Metric),
         Metric = if_else(Metric == "excretionRateN",
                          "Ammonium Excretion Rate", Metric),
         Metric = if_else(Metric == "excretionRateP",
                          "Phosphate Excretion Rate", Metric)) %>% 
  mutate(Metric = factor(Metric, levels = c("Abundance","Carbon Weight",
                                            "Respiration Rate",
                                            "Ammonium Excretion Rate",
                                            "Phosphate Excretion Rate",
                                            "Clearance Rate")))


# Cowplot version
metrics <- levels(zoop.163.ctotal.plot$Metric)
plots <- vector("list",2)
count <- 1
for (i in c(1:6)) {
   plots[[count]] <- ggplot(zoop.163.ctotal.plot %>% 
                                  filter(Metric == metrics[i]), 
                                aes(x = Bioregion, y = Value.mean)) +
      geom_bar(aes(fill = fungrp), position = "stack", stat = "identity") +
      scale_fill_manual(values = rev(fungrp.colors), 
                        name = "Functional\nGroup") +
      scale_y_continuous(expand = c(0, 0),
                         name =unique(metrics[i])) +
      theme(legend.position = "none",
            axis.title.y.right = element_markdown()) 
  
  count <- count + 1
}

# Combine the plots into a grid
plot_grid_layout <- plot_grid(plotlist = plots, ncol = 3, labels = "auto") 
plot_grid_layout


# Seems like get_legend() does not work anymore for the updated ggplot version3.5 .. have to manually annotate the legend for now
# legend_b1 <- get_legend(
#   plots[[1]]+
#     theme(legend.position = "bottom") )
# final_plot <- plot_grid(
#   plot_grid_layout, legend_b1, 
#   ncol = 1, 
#   rel_heights = c(3, 0.3) 
# )

ggsave(filename = here("figures/fig_4_com_total_rates_Aug2025.png"),
       width = 12, height= 6, units = "in", dpi = 600, bg = "white")

# For manual legend annotation
ggplot(zoop.163.ctotal.plot,
  aes(x = Bioregion, y = Value.mean)) +
  geom_bar(aes(fill = fungrp), position = "stack", stat="identity") +
  scale_fill_manual(values = rev(fungrp.colors),
           name = "Functional Group",
           labels = paste0("FG", unique(zoop.163.ctotal.plot$fungrp))) +
  scale_y_continuous(expand = c(0, 0),
                     name = "Community Total Value") +
  theme(legend.position = "bottom",  
        legend.box = "horizontal",
        axis.title.y.right = element_markdown()) +
  guides(fill = guide_legend(nrow = 1, 
                             size = 12,
                            label.position = "right",
                            byrow = TRUE)) +
  facet_wrap(~Metric, scales = "free_y")


ggsave(filename = here("figures/fig_4_com_total_rates_legend_Aug2025.png"),
       width = 12, height= 10, units = "in", dpi = 600, bg = "white")

```

# 4. Table1 Bioregion comparison

## Sig test CWM 
```{r}
Trait.labs <- c("Body length","Dry weight","Water content",
               "Carbon weight",
               "Nitrogen weight","Phosphorus weight",
               "Carbon content","Nitrogen content","Phosphorus content",
               "C:N","C:P","N:P",
               "Ind. respiration rate","Ws. respiration rate",
               "Ind. ammonium\nexcretion rate","Ws. ammonium\nexcretion rate",
               "Ind. phosphate\nexcretion rate",
               "Ws. phosphate\nexcretion rate",
               "Ind. clearance rate", "Ws. clearance rate",
               "Prop. current feeders","Prop. cruise feeders",
               "Prop. active\nambush feeders","Prop. passive\nambush feeders",
               "Prop. herbivores","Prop. omnivores","Prop. carnivores",
               "Prop. detritivores",
               "Prop. broadcast\nspawning","Prop. brood\nspawning")
names(Trait.labs) <- c("bodyLength","dryWeight","waterPWW",
               "carbonWeight","nitrogenWeight","phosphorusWeight",
               "carbonPDW","nitrogenPDW","phosphorusPDW","ratioCN","ratioCP","ratioNP",
               "respirationRate","respirationRate_WSC",
               "excretionRateN","excretionRateN_WSDW",
               "excretionRateP", "excretionRateP_WSDW",
               "clearanceRate","clearanceRate_WSC",
               "FM.current","FM.cruise","FM.active.ambush","FM.passive.ambush",
               "TG.herbivore","TG.omnivore","TG.carnivore","TG.detritivore",
               "RM.broadcasting","RM.brooding")

cwm.kw <- zoop.163.cwm %>% 
  group_by(Trait) %>% 
  do(tidy(kruskal.test(x=.$CWM, g =.$Bioregion)))

cwm.dunn <- zoop.163.cwm %>% 
  group_by(Trait) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$CWM, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted >= 0.0005) %>% 
  arrange(match(Trait, names(Trait.labs)))


cwm.sum <- zoop.163.cwm %>% 
  group_by(Bioregion, Trait) %>% 
  summarise(ValMean = mean(CWM),
            ValSD = sd(CWM), 
            ValMedian = median(CWM),
            ValIQR = IQR(CWM),
            ValMin = min(CWM),
            ValMax = max(CWM),
            ValN = n(),
            .groups = "drop") %>% 
  arrange(Bioregion, Trait) 


# Per bioregion and functional group
cwm.table <- cwm.sum %>% 
  select(Bioregion, Trait, ValMean, ValSD) %>% 
  mutate(ValText = if_else(Trait %in% c("detritivore","phosphorusWeight"), 
           paste(round(ValMean,3),"\u00b1",round(ValSD,3)),
           paste(round(ValMean,2),"\u00b1",round(ValSD,2)))) %>% 
  select(-c(ValMean, ValSD)) %>% 
  pivot_wider(names_from = Bioregion,
              values_from = ValText) %>% 
  arrange(match(Trait, names(Trait.labs)))
cwm.table$Trait <- Trait.labs

write.csv(cwm.table,
          file = here("tables/cwm_bioregions_to_annotate.csv"), row.names = FALSE)
```

## Sig test Ctotals
```{r}
ctotal.kw <- zoop.163.ctotal %>% 
  group_by(Trait) %>% 
  do(tidy(kruskal.test(x=.$CTotal, g =.$Bioregion)))

ctotal.dunn <- zoop.163.ctotal %>% 
  group_by(Trait) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$CTotal, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted >= 0.0005) 


ctotal.sum <- zoop.163.ctotal %>% 
  group_by(Bioregion, Trait) %>% 
  summarise(ValMean = mean(CTotal),
            ValSD = sd(CTotal), 
            ValMedian = median(CTotal),
            ValIQR = IQR(CTotal),
            ValMin = min(CTotal),
            ValMax = max(CTotal),
            ValN = n(),
            .groups = "drop") %>% 
  arrange(Bioregion, Trait) 


# Per bioregion and functional group
ctotal.table <- ctotal.sum %>% 
  select(Bioregion, Trait, ValMean, ValSD) %>% 
  mutate(ValText = paste(round(ValMean),"\u00b1",round(ValSD))) %>%
  select(-c(ValMean, ValSD)) %>% 
  pivot_wider(names_from = Bioregion,
              values_from = ValText) %>% 
  arrange(match(Trait, names(Trait.labs)))

# write.csv(ctotal.table,
#           file = here("tables/ctoal_bioregions_to_annotate.csv"), row.names = FALSE)
```

# 5. Fig 5. Trait comparison of bioregions
## Rough figure using facet wrap
```{r, eval = FALSE}
traits.meansd <- zoop.163.cwm %>% 
  group_by(Bioregion, Trait) %>% 
   summarise(ValMean = mean(CWM),
            ValSD = sd(CWM), 
            ValN = n(),
            .groups = "drop") %>% 
  arrange(Bioregion, Trait) %>% 
  mutate(Bioregion2 = if_else(Bioregion == "Offshore","OS", "DF"),
         Bioregion2 = if_else(Bioregion == "DeepShelf","DS", Bioregion2),
         Bioregion2 = if_else(Bioregion == "Nearshore","NS", Bioregion2)) %>% 
  mutate(Bioregion2 = factor(Bioregion2, levels = c("OS","DS","NS","DF"))) %>% 
  mutate(ValSE = ValSD / sqrt(ValN))

ggplot(traits.meansd, aes(x = Bioregion, y = ValMean, fill = Bioregion)) +
  # geom_errorbar(aes(ymin = ValMean - ValSE, ymax = ValMean + ValSE),
  geom_errorbar(aes(ymin = ValMean - ValSD, ymax = ValMean + ValSD,
                    color = Bioregion),
                 width = 0, size = 1) +
  geom_point(aes(fill = Bioregion), shape = 21, size = 5) +
  scale_color_manual(values = clrs.bioregion) +
  scale_fill_manual(values = clrs.bioregion) +
  theme(legend.position = "bottom",
        axis.title = element_blank(),
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  facet_wrap(~Trait, scale = "free_y", ncol = 6)
```


## Figure with proper axis labels based on the table of results in Table S3.
```{r, fig.width=12}
bioregion.table <- readxl::read_xlsx(here("tables/Table_bioregion_comparisons.xlsx")) 

bioreg.mean <- bioregion.table %>% 
  select(ID, Trait, Label, Plotgrid_label, OS_MEAN, DS_MEAN, NS_MEAN, DF_MEAN) %>% 
  pivot_longer(cols = c("OS_MEAN","DS_MEAN","NS_MEAN","DF_MEAN"),
               names_to = "Bioregion", values_to = "meanval") %>% 
  mutate(Bioregion = str_remove(Bioregion, "_MEAN"))

bioreg.sd <- bioregion.table %>% 
  select(ID, Trait, OS_SD, DS_SD, NS_SD, DF_SD) %>% 
  pivot_longer(cols = c("OS_SD","DS_SD","NS_SD","DF_SD"),
               names_to = "Bioregion", values_to = "sdval") %>% 
  mutate(Bioregion = str_remove(Bioregion, "_SD"))

bioregion.table.2 <- left_join(bioreg.mean, bioreg.sd,
                               by = join_by(ID, Trait, Bioregion)) %>% 
  mutate(Trait = factor(Trait, levels = unique(bioregion.table$Trait))) %>% 
  
  mutate(Bioregion2 = if_else(Bioregion == "OS","Offshore", "Deep Fjord"),
         Bioregion2 = if_else(Bioregion == "DS","Deep Shelf", Bioregion2),
         Bioregion2 = if_else(Bioregion == "NS","Nearshore", Bioregion2)) %>% 
  mutate(Bioregion = Bioregion2) %>% 
  mutate(Bioregion = factor(Bioregion, levels = c("Offshore","Deep Shelf",
                                                  "Nearshore","Deep Fjord"))) 


metrics <- unique(bioregion.table$Trait)
labels <- unique(bioregion.table$Label)
plots <- vector("list",2)
count <- 1
for (i in c(1:35)) {
  plots[[count]] <- ggplot(bioregion.table.2 %>% 
                             filter(Trait == metrics[i]), 
                           aes(x = Bioregion, y = meanval, fill = Bioregion)) +
    geom_errorbar(aes(ymin = meanval - sdval, ymax = meanval + sdval,
                      color = Bioregion),
                  width = 0, size = 1) +
    geom_point(aes(fill = Bioregion), shape = 21, size = 5) +
    scale_color_manual(values = clrs.bioregion) +
    scale_fill_manual(values = clrs.bioregion) +
    theme(legend.position = "none",
          # axis.title.x = element_blank(),
          axis.title.y = element_markdown(vjust = 0),
          axis.text.y = element_text(size = 14),
          axis.ticks.x = element_blank(),
          axis.text.x = element_blank()) +
    ylab(labels[i]) + xlab("")
  
  count <- count + 1
}

# Combine the plots into a grid
plot_grid_layout <- plot_grid(plotlist = plots, ncol = 6, 
                              align = "hv", labels = bioregion.table$Plotgrid_label) 
plot_grid_layout


ggsave(filename = here("figures/fig_7_bioregion_comparison.png"),
    width = 12, height= 14, units = "in", dpi = 600, bg = "white")
```
