---
title: "Functional group analysis"
author: "Patrick Pata"
date: "2023-10-24"
output: html_document
---

For description of diversity indices and figures see: https://cmlmagneville.github.io/mFD/articles/mFD_general_workflow.html#test-correlation-between-functional-axes-and-traits 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Prepare data
## Load libraries and data
```{r}
library(tidyverse)
library(here)
library(gawdis)
library(data.table) # for calculating convex hulls
library(ggrepel)
library(ggtext)
library(broom)
set_here(path = "..")
`%notin%` <- Negate(`%in%`)


theme_set(theme_bw())

load(here("data/BC_species_trait_matrices_20240604.RData"))
rm(allomModels)


# Assign colors to major groups
majorgroup.colors <- data.frame(
  majorgroup = c("Calanoid","Non-calanoid","Amphipod", "Mysid",
                 "Euphausiid", "Ostracod","Polychaete","Pteropod",
                 "Chaetognath","Appendicularian","Thaliacean",
                 "Hydromedusae","Siphonophore","Ctenophore"),
  color = c("blue4","dodgerblue","slateblue1", "seagreen3",
                   "cyan3","forestgreen","lightpink2","green2",
                   "darkorchid4","violetred","tomato1",
                   "tan1","yellow2","red3"))
                   

fungrp.colors <- c("olivedrab3","darkolivegreen4","darkgreen","gold3",
                  "lightcoral","magenta4","orchid",
                  "darkorange2","steelblue2","turquoise3")
                               
regions <- c("Offshore","DeepShelf","Nearshore","DeepFjord")
clrs.bioregion <- c("#004DFB","#FB20CA","#FFA500","#20FF00")

```

## Transform trait matrix
```{r,eval=TRUE}
# 1. select traits for FG analysis 
# 2. transform necessary traits
# 3. combine with binary data
# 4. filter species based on presence (run both and see if grouping is similar)
# 5. setup groups and fuzzy variables

# # List of species to include
species.include.list <- zoop.bc.list %>%
  # Exclude if rare
  filter(presence.perc >= 1) %>%
  #  Exclude because of high uncertainty in trait values. Especially for all polychaetes outside family Tomopteridae.
  filter(Species %notin% c("Thalassocalyce inconstans","Corolla spectabilis")) %>%
  filter(!(majorgroup == "Polychaete" & !grepl("Tomopteris",Species)))
  
  
species.traits <- lvl3.num.matrix %>% 
  select(Species, dryWeight, waterPWW, 
         respirationRate_WSC, clearanceRate_WSC,
         excretionRateN_WSDW,
         ) %>% 
  # log-transform some continuous trait values
  mutate(dryWeight = log(dryWeight),
         respirationRate_WSC = log(respirationRate_WSC),
         excretionRateN_WSDW = log(excretionRateN_WSDW),
         clearanceRate_WSC = log(clearanceRate_WSC)
         ) %>% 
  # Remove species with high uncertainty in trait estimates + rare species
  # filter(Species %in% filter(zoop.bc.list, presence.perc >= 1)$Species) %>% 
  filter(Species %in% species.include.list$Species) %>%
  # combine with categorical data as fuzzy variables
  left_join(lvl3.cat.matrix, by = "Species") %>% 
  # select(-c(RM.broadcasting, RM.brooding)) %>%
  column_to_rownames("Species") 

# Identify species with incomplete trait information
ii <- which(!is.na(rowMeans(species.traits)))
# Analyze all species with complete trait information
species.traits <- species.traits[ii,]

trait.list <- data.frame(
  traitName = c("dryWeight","waterPWW", 
                "respirationRate_WSC", "clearanceRate_WSC",
                "excretionRateN_WSDW", 
                "RM.broadcasting","RM.brooding",
                "FM.cruise","FM.current","FM.active.ambush","FM.passive.ambush",
                "TG.herbivore","TG.omnivore","TG.carnivore","TG.detritivore")) %>% 
  mutate(trait.group = traitName) %>% 
  mutate(trait.group = if_else(grepl("FM.",trait.group),
                               "feedingMode",trait.group),
         trait.group = if_else(grepl("RM.",trait.group),
                               "reproductionMode",trait.group),
         trait.group = if_else(grepl("TG.",trait.group),
                               "trophicGroup",trait.group))


AA <- lvl3.numeric %>% 
  filter(traitName %in% c("waterPWW","dryWeight",
                "respirationRate_WSC", "clearanceRate_WSC",
                "excretionRateN_WSDW")) %>% 
  # distinct(Species, majorgroup, presence.perc) %>% 
  filter(presence.perc > 1) %>% 
  filter(valueConf >= 4) %>% 
  group_by(Species, majorgroup, presence.perc) %>% 
  summarise(ntraits = n())

```


## Identify colinearity between traits
```{r, eval=FALSE, fig.height=5, fig.width=5}
# Prepare a list of pairs of traits to correlate
trait.cor.list <- trait.list[1:13,]
trait.cor <- data.frame(trait1 = character(), trait2 = character())
for (i in c(1:(nrow(trait.cor.list)-1))) {
  for (j in c(i:nrow(trait.cor.list))) {
    trait.cor <- trait.cor %>%
      bind_rows(data.frame(trait1 = trait.cor.list$traitName[i],
                           trait2 = trait.cor.list$traitName[j]))
  }
}
trait.cor <- trait.cor %>%
  filter(trait1 != trait2)

# Loop through correlations
cor.tests <- data.frame()
for (i in c(1:nrow(trait.cor))){ #
  AA <- species.traits %>%
    rownames_to_column("Species") %>% 
    pivot_longer(cols = -Species, names_to = "traitName", values_to = "traitValue") %>% 
    filter(!is.na(traitValue)) %>% 
    filter(traitName %in% c(trait.cor$trait1[i], trait.cor$trait2[i])) %>%
    select(Species, traitName, traitValue) %>%
    pivot_wider(names_from = traitName, values_from = traitValue) %>%
    filter(!is.na(get(trait.cor$trait1[i])), !is.na(get(trait.cor$trait2[i]))) %>%
    column_to_rownames("Species")

  res <- data.frame(trait1 = colnames(AA[1]), trait2 = colnames(AA[2]),
                    nspecies = nrow(AA), r = cor(AA)[1,2])

  cor.tests <- cor.tests %>%
    bind_rows(res)
}


# DIY correlation plot
cor.tests <- cor.tests %>% 
  rownames_to_column("analysis")
cor.table <- cor.tests %>% 
  bind_rows(dplyr::select(cor.tests, trait2 = trait1, trait1 = trait2, 
                          everything())) %>% 
   mutate(trait1 = factor(trait1, levels = trait.cor.list$traitName),
         trait2 = factor(trait2, levels = trait.cor.list$traitName)) %>% 
  mutate(trait1.num = as.numeric(trait1),
         trait2.num = as.numeric(trait2)) %>% 
  # retain only one side of the matrix
  arrange(trait1, trait2) %>% 
  group_by(analysis) %>% 
  filter(row_number() == 1) %>% 
# How many correlation coefficients >0.50 and nspecies > 100
  mutate(r.mag = if_else(abs(r) > 0.5 & nspecies >= 0,
         "s","w"))

ggplot() +
  # Correlation data
  geom_point(data = cor.table, aes(x = trait1.num, y = trait2.num, color = r.mag,
                        fill = r),
             shape = 21, size = 12) +
  geom_text(data = cor.table, aes(x = trait1.num, y = trait2.num, 
                                  label = sprintf("%0.2f", round(r, digits = 2)))) +
  # scale_color_gradient2(low = "blue3", mid = "white", high = "red2", 
  #                      limits = c(-1,1), name = "corr coef") +
  scale_color_manual(values = c("black",NA), guide = "none") +
  scale_fill_gradient2(low = "blue3", mid = "white", high = "red2", 
                       limits = c(-1,1), name = "r") +
  
  scale_x_discrete(limits = trait.cor.list$traitName) +
  scale_y_reverse(breaks = 1:nrow(trait.cor.list), labels = trait.cor.list$traitName,
                  limits = c(nrow(trait.cor.list)+0.5,0.5), expand = c(0, 0))+
  scale_size_continuous(breaks = c(10,100,200,250), name = "N") +
  # scale_y_discrete(limits = (trait.cor.list$traitName))  +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45,  hjust=1),
        axis.title = element_blank(),
        aspect.ratio = 1,
        legend.position = "right",
        legend.box = "vertical",
        legend.key.height = unit(0.5,"in"),
        legend.key.width=unit(0.3,"in"),
        text = element_text(size = 16),
        panel.grid.minor.y = element_blank(),
        panel.background = element_rect(fill='white'),
        plot.background = element_rect(fill='white', color=NA))

# ggsave(filename = here("figures/BC_traits_correlation.png"),
#     width = 10, height= 8, units = "in", dpi = 300)

rm(trait.cor.list, trait.cor, cor.tests, cor.table, res, AA)
```

# 2. Trait dissimilarity between species

The calculation of the dissimilarity between species according to trait values should be sensitive to the value type (numeric, binary, fuzzy coded categorical) and the relationship between similar traits (e.g., size: length, weight).

## Gawdis optimized
```{r, fig.width=6}
set.seed(888)
gaw.dist <- gawdis::gawdis(species.traits,
                           w.type = "analytic", #use "optimized" if with NAs
                           groups = trait.list$trait.group,
                           fuzzy = c("feedingMode","trophicGroup",
                                     "reproductionMode"),
                           opti.maxiter = 1000)


# Correlation between traits
cor.mat <- attr(gaw.dist, "cor.mat")
cormat.pval <- ggcorrplot::cor_pmat(cor.mat)
ggcorrplot::ggcorrplot(cor.mat, hc.order = TRUE, 
     outline.col = "white", p.mat = cormat.pval) #  insig = "blank")

# ggsave(filename = here("figures/Gawdis_species_cormat.png"),
#     width = 13, height= 12, units = "in", dpi = 300, bg = "white")


gaw.stats <- data.frame(correls = attr(gaw.dist, "correls"),
                        weights = attr(gaw.dist, "weights")) %>% 
  rownames_to_column("traitName")
# write.csv(gaw.stats, file = here("tables/gawdis_correls_weights.csv"),
#           row.names = FALSE)
```

## Equal weights
The equal weights gawdis tends to overestimate dissimilarity between species but the r statistic is quite high at r = 0.95.
```{r, eval=FALSE}
# # Apply weights within groups
# Equal weights between all groups
# # Note that groups.weight = TRUE can't be calculated when there are fuzzy variables.
gawdis.equal <- gawdis::gawdis(species.traits,
                               w.type = "equal",
                               groups = trait.list$trait.group,
                               # groups.weight = TRUE,
                               fuzzy = c("feedingMode","trophicGroup",
                                         "reproductionMode"))

plot(gaw.dist, gawdis.equal)
abline(0,1)

mantel(gaw.dist, gawdis.equal)
```

# 3. Ordination of gaw.dist with mFD package
MAD/absolute: absolute difference in distances
RMSE/squared: more weight to large deviation
Scaled: scaling distances in functional space to have the same maximum

*NOTE* That the PCoA from the quality.fspaces() function is the same as calculating pcoa(gaw.dist, correction = "cailliez") 

All 4 metrics determine 6 axes as the best representation of the trait space.
```{r, fig.width=6}
gaw.space <- mFD::quality.fspaces(sp_dist = gaw.dist,
                             fdendro = "average",
                             maxdim_pcoa = 10,
                             deviation_weighting = c("absolute", "squared"),
                             fdist_scaling = c(TRUE, FALSE))

# The quality of the dimension reduction in relation to retaining the original trait dissimilarity
gaw.space$quality_fspaces
# retrieve the functional space associated with minimal quality metric: 
apply(gaw.space$quality_fspaces, 2, which.min)

# Plot quality metrics
gaw.space$"quality_fspaces" %>%
  tibble::as_tibble(rownames = "Funct.space") %>%
  tidyr::pivot_longer(cols =! Funct.space, names_to = "quality_metric", values_to = "Quality") %>%
  ggplot2::ggplot(ggplot2::aes(x = Funct.space, y = Quality, 
                               color = quality_metric, shape = quality_metric)) +
  ggplot2::geom_point() 

mFD::quality.fspaces.plot(
  fspaces_quality = gaw.space, 
  quality_metric  = "mad",
  fspaces_plot    = c("tree_average", "pcoa_2d", "pcoa_5d",'pcoa_9d'))

# ggsave(filename = here("figures/Quality_functional_space_reduction.png"),
#     width = 13, height= 10, units = "in", dpi = 300, bg = "white")
```

## Save gawdis analysis results
```{r}
save(gaw.space, gaw.dist, species.include.list,
  file = here("data/fundiv_gawdis_results_20240729.RData")
)
```


# 4. Coordinate-based functional diversity indices
## Load data
```{r}
# load(here("data/fundiv_gawdis_results_20240729.RData"))

# Community data
load(here("data/key_fg_abundance_biomass_20240626.RData"))

# Functional group data + traits
load(here("data/functional_grouping_results_20240617.RData"))
```

## Relative abundance and biomass of each functional group between bioregions
Note that this chunk takes a while to compute.
```{r, fig.width=5}
species.pcoa <- gaw.space$details_fspaces$sp_pc_coord[,c(1:5)]

strip_fg <- ggh4x::strip_themed(background_x = ggh4x::elem_list_rect(
  fill = gplots::col2hex(fungrp.colors)))

# Create sites x functional group matrix
bc.com.mat.abun <- zoop.bc.species.fun %>% 
  select(Key, Species, Abundance.m2) %>%
  # select(Key, Species, Biomass.mg.m2) %>%
  pivot_wider(names_from = Species, values_from = Abundance.m2,
              values_fill = 0) %>% 
  column_to_rownames("Key")

bc.com.mat.biom <- zoop.bc.species.fun %>% 
  select(Key, Species, Biomass.mg.m2) %>%
  # select(Key, Species, Biomass.mg.m2) %>%
  pivot_wider(names_from = Species, values_from = Biomass.mg.m2,
              values_fill = 0) %>% 
  column_to_rownames("Key")


fundiv.abun <- mFD::alpha.fd.multidim(sp_faxes_coord = species.pcoa,
                       asb_sp_w = as.matrix(bc.com.mat.abun), verbose = FALSE)

fundiv.biom <- mFD::alpha.fd.multidim(sp_faxes_coord = species.pcoa,
                       asb_sp_w = as.matrix(bc.com.mat.biom), verbose = FALSE)


# Taxonomic diversity + by functional group
taxdiv <-  zoop.bc.species.fun %>% 
  select(Key, Species, fungrp, Abundance.m2, Biomass.mg.m2) %>% 
  group_by(Key) %>% 
  mutate(Richness = specnumber(Abundance.m2),
         Diversity.Shannon = diversity(Abundance.m2, index = "shannon"),
         Diversity.Simpson = diversity(Abundance.m2, index = "simpson")) %>% 
  mutate(Evenness = Diversity.Shannon/log(Richness)) %>% 
  ungroup() %>% 
  group_by(Key, fungrp) %>% 
  mutate(Richness.fungrp = n()) %>% 
  ungroup() %>% 
  select(-c(Species, Abundance.m2, Biomass.mg.m2)) %>% 
  distinct()
  
```

## Distribution of functional diversity metrics between bioregions
```{r}
Metric.labs <- c("Species Richness","Functional Richness",
                 "Functional Dispersion","Functional Evenness",
                 "Functional Divergence","Functional Originality")
names(Metric.labs) <- c("Sp.Ric","FRic","FDis","FEve","FDiv","FOri")

zoopmeta <- zoop.bc.species.fun %>% 
  distinct(Key, Bioregion)

fundiv.ind.biom <- fundiv.biom$functional_diversity_indices %>% 
  rownames_to_column("Key") %>% 
  select(Key, Sp.Ric = sp_richn, FRic = fric, FDis = fdis, FEve = feve, 
         FDiv = fdiv, FOri = fori)  %>% 
  left_join(select(zoopmeta, Key, Bioregion), by = "Key") %>% 
  relocate(Bioregion) %>% 
  column_to_rownames("Key")


fundiv.ind.biom.long <- fundiv.ind.biom %>% 
  rownames_to_column("Key") %>% 
  pivot_longer(cols = -c(Key,Bioregion), names_to = "Metric", values_to = "Value") %>% 
  mutate(Metric = factor(Metric, 
                         levels = c("Sp.Ric","FRic","FDis","FEve",
                                    "FDiv","FOri"))) 

# KW and Dunn test
fundiv.ind.kw <- fundiv.ind.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(tidy(kruskal.test(x=.$value, g =.$Bioregion)))
fundiv.ind.dun <- fundiv.ind.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$value, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted > 0.0005)


save(fundiv.abun, fundiv.biom, taxdiv, fundiv.ind.biom, fundiv.ind.biom.long,
fundiv.ind.kw, fundiv.ind.dun,
     file = here("data/fundiv_3721_163species_20240618.RData"))
```



# Plot means as bar plots for summary figure
```{r}
fundiv.mean <- fundiv.ind.biom.long %>% 
  filter(Metric != "Sp.Ric") %>% 
  group_by(Bioregion, Metric) %>% 
  summarise(Value = mean(Value), .groups = "drop")

ggplot(fundiv.mean, aes(y = Metric, x = Value)) +
  geom_bar(stat = "identity", aes(fill = Bioregion), alpha = 1) +
  scale_x_continuous(limits = c(0,1), expand = c(0, 0),
                     breaks = c(0,0.25,0.5,0.75,1),
                     labels = c("0","0.25","0.5","0.75","1")) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = clrs.bioregion) +
  theme(legend.position = "none") +
  facet_wrap(~Bioregion, nrow = 1) 

ggsave(file = here("figures/fig9/fig9_fundiv_mean.png"),
       width = 8, height = 1.8, dpi = 300 )
```







