---
title: "Functional group analysis"
author: "Patrick Pata"
date: "2023-10-24"
output: html_document
---

This file contains the scripts for calculating the functional diversity indices and reporting the between-bioregion differences. The calculated diversity indices are provided in the data loaded. Recalculating the diversity indices is possible with a raw species level community data. Note that mapping the functional diversity indices requires some mapping libraries and bathymetry data.

For description of diversity indices and figures see:https://cmlmagneville.github.io/mFD/articles/mFD_general_workflow.html#about-this-tutorial

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load libraries and data
```{r}
library(tidyverse)
library(here)
library(gawdis)
library(data.table) # for calculating convex hulls
library(ggrepel)
library(ggtext)
library(broom)

theme_set(theme_bw())
`%notin%` <- Negate(`%in%`)
set_here(path = "..")

# Only need zoopmeta dataframe from regionalization data
load(here("data/BC_bioregionalization_metadata.RData"))
rm(zoopenvi)

# Functional diversity indices calculated in the chunk below
load(here("data/fundiv_3721_163species_20240618.RData"))

# Assign colors to major groups
majorgroup.colors <- data.frame(
  majorgroup = c("Calanoid","Non-calanoid","Amphipod", "Mysid",
                 "Euphausiid", "Ostracod","Polychaete","Pteropod",
                 "Chaetognath","Appendicularian","Thaliacean",
                 "Hydromedusae","Siphonophore","Ctenophore"),
  color = c("blue4","dodgerblue","slateblue1", "seagreen3",
                   "cyan3","forestgreen","lightpink2","green2",
                   "darkorchid4","violetred","tomato1",
                   "tan1","yellow2","red3"))
                   

fungrp.colors <- c("olivedrab3","darkolivegreen4","darkgreen","gold3",
                  "lightcoral","magenta4","orchid",
                  "darkorange2","steelblue2","turquoise3")
                               
regions <- c("Offshore","DeepShelf","Nearshore","DeepFjord")
clrs.bioregion <- c("#004DFB","#FB20CA","#FFA500","#20FF00")

```


# 2. Recalculate functional diversity indices
Note that this chunk takes a while to compute. If recalculating, make sure to uncomment the save line and update the file storing the functional diversity results.
```{r, fig.width=5, eval=FALSE}
# Community data
load(here("data/restricted_data/key_fg_abundance_biomass_20240626.RData"))

# Gawdis results from file 4
load(here("data/fundiv_gawdis_results_20240729.RData"))

# Functional group data + traits
load(here("data/functional_grouping_results_20240617.RData"))

species.pcoa <- gaw.space$details_fspaces$sp_pc_coord[,c(1:5)]

strip_fg <- ggh4x::strip_themed(background_x = ggh4x::elem_list_rect(
  fill = gplots::col2hex(fungrp.colors)))

# Create sites x functional group matrix
bc.com.mat.abun <- zoop.bc.species.fun %>% 
  select(Key, Species, Abundance.m2) %>%
  # select(Key, Species, Biomass.mg.m2) %>%
  pivot_wider(names_from = Species, values_from = Abundance.m2,
              values_fill = 0) %>% 
  column_to_rownames("Key")

bc.com.mat.biom <- zoop.bc.species.fun %>% 
  select(Key, Species, Biomass.mg.m2) %>%
  # select(Key, Species, Biomass.mg.m2) %>%
  pivot_wider(names_from = Species, values_from = Biomass.mg.m2,
              values_fill = 0) %>% 
  column_to_rownames("Key")


fundiv.abun <- mFD::alpha.fd.multidim(sp_faxes_coord = species.pcoa,
                       asb_sp_w = as.matrix(bc.com.mat.abun), verbose = FALSE)

fundiv.biom <- mFD::alpha.fd.multidim(sp_faxes_coord = species.pcoa,
                       asb_sp_w = as.matrix(bc.com.mat.biom), verbose = FALSE)


# Taxonomic diversity + by functional group
taxdiv <-  zoop.bc.species.fun %>% 
  select(Key, Species, fungrp, Abundance.m2, Biomass.mg.m2) %>% 
  group_by(Key) %>% 
  mutate(Richness = specnumber(Abundance.m2),
         Diversity.Shannon = diversity(Abundance.m2, index = "shannon"),
         Diversity.Simpson = diversity(Abundance.m2, index = "simpson")) %>% 
  mutate(Evenness = Diversity.Shannon/log(Richness)) %>% 
  ungroup() %>% 
  group_by(Key, fungrp) %>% 
  mutate(Richness.fungrp = n()) %>% 
  ungroup() %>% 
  select(-c(Species, Abundance.m2, Biomass.mg.m2)) %>% 
  distinct()
  

# Distribution of functional diversity metrics between bioregions

Metric.labs <- c("Species Richness","Functional Richness",
                 "Functional Dispersion","Functional Evenness",
                 "Functional Divergence","Functional Originality")
names(Metric.labs) <- c("Sp.Ric","FRic","FDis","FEve","FDiv","FOri")

zoopmeta <- zoop.bc.species.fun %>% 
  distinct(Key, Bioregion)

fundiv.ind.biom <- fundiv.biom$functional_diversity_indices %>% 
  rownames_to_column("Key") %>% 
  select(Key, Sp.Ric = sp_richn, FRic = fric, FDis = fdis, FEve = feve, 
         FDiv = fdiv, FOri = fori)  %>% 
  left_join(select(zoopmeta, Key, Bioregion), by = "Key") %>% 
  relocate(Bioregion) %>% 
  column_to_rownames("Key")


fundiv.ind.biom.long <- fundiv.ind.biom %>% 
  rownames_to_column("Key") %>% 
  pivot_longer(cols = -c(Key,Bioregion), names_to = "Metric", values_to = "Value") %>% 
  mutate(Metric = factor(Metric, 
                         levels = c("Sp.Ric","FRic","FDis","FEve",
                                    "FDiv","FOri"))) 

# KW and Dunn test
fundiv.ind.kw <- fundiv.ind.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(tidy(kruskal.test(x=.$value, g =.$Bioregion)))
fundiv.ind.dun <- fundiv.ind.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$value, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted > 0.0005)


# save(fundiv.abun, fundiv.biom, taxdiv, fundiv.ind.biom, fundiv.ind.biom.long,
# fundiv.ind.kw, fundiv.ind.dun,
#      file = here("data/fundiv_3721_163species_20240618.RData"))
```

# 3. Fig S5 Functional diversity plot
## Distribution of functional diversity metrics between bioregions
```{r}
Metric.labs <- c("Species Richness","Functional Richness",
                 "Functional Dispersion","Functional Evenness",
                 "Functional Divergence","Functional Originality")
names(Metric.labs) <- c("Sp.Ric","FRic","FDis","FEve","FDiv","FOri")


fdivplots <- vector("list",2)
count <- 1
for (i in c(1:6)) {
   fdivplots[[count]] <- ggplot(filter(fundiv.ind.biom.long, 
                                        Metric == names(Metric.labs)[i]),
                      aes(x = Bioregion, y = Value, fill = Bioregion)) +
  geom_violin(aes(color = Bioregion), alpha = 0.6) + 
  geom_boxplot(width = 0.3, color = "grey25", outlier.shape = NA) +
  xlab("") +
  ylab(Metric.labs[i]) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme(legend.position = "none",  
        axis.title = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.ticks.x = element_blank(),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        strip.text.x = element_text(size = 12)) +
  scale_fill_manual(values = clrs.bioregion) +
  scale_color_manual(values = clrs.bioregion) 
  
  count <- count + 1
}

legend_b1 <- get_legend(
   fdivplots[[1]] + 
    # guides(color = guide_legend(nrow = 2)) +
    theme(legend.position = "bottom")
)


plot.violin <- plot_grid(
  plot_grid(plotlist = fdivplots, nrow = 1, labels = c("c","d","e","f","g","h")),
  legend_b1, 
  ncol = 1, rel_heights = c(1, .1))

# ggsave(file = here("figures/fig_7_fundiv_violin.png"),
#        height = 4, width = 16, dpi = 300, bg = "white")
```


# 4. Fig 7 Plot means as bar plots for summary figure
```{r}
fundiv.mean <- fundiv.ind.biom.long %>% 
  filter(Metric != "Sp.Ric") %>% 
  group_by(Bioregion, Metric) %>% 
  summarise(Value = mean(Value), .groups = "drop")

ggplot(fundiv.mean, aes(y = Metric, x = Value)) +
  geom_bar(stat = "identity", aes(fill = Bioregion), alpha = 1) +
  scale_x_continuous(limits = c(0,1), expand = c(0, 0),
                     breaks = c(0,0.25,0.5,0.75,1),
                     labels = c("0","0.25","0.5","0.75","1")) +
  scale_y_discrete(limits = rev) +
  scale_fill_manual(values = clrs.bioregion) +
  theme(legend.position = "none") +
  facet_wrap(~Bioregion, nrow = 1) 

ggsave(file = here("figures/fig9_fundiv_mean.png"),
       width = 8, height = 1.8, dpi = 300 )
```

# 5. Table 1 Significance test between bioregions
```{r}
# KW and Dunn test
fundiv.ind.kw <- fundiv.ind.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(tidy(kruskal.test(x=.$value, g =.$Bioregion)))
fundiv.ind.dun <- fundiv.ind.biom %>% 
  gather(key, value, -Bioregion) %>% 
  group_by(key) %>% 
  do(as.data.frame(dunn.test::dunn.test(x=.$value, g =.$Bioregion, 
                                        method = "bh", alpha = 0.001))) %>% 
  filter(P.adjusted > 0.0005)


# Output the means +- sd
fdiv.table <- fundiv.ind.biom.long %>%
  group_by(Bioregion, Metric) %>% 
  summarise(ValMean = mean(Value),
            ValSD = sd(Value), 
            ValMedian = median(Value),
            ValIQR = IQR(Value),
            ValMin = min(Value),
            ValMax = max(Value),
            ValN = n(),
            .groups = "drop") %>% 
  arrange(Bioregion, Metric) %>% 
  
  select(Bioregion, Metric, Bioregion, ValMean, ValSD) %>% 
  mutate(ValText = paste(round(ValMean,2),"\u00b1",round(ValSD,2))) %>% 
  select(-c(ValMean, ValSD)) %>% 
  pivot_wider(names_from = Bioregion,
              values_from = ValText) 

# write.csv(fdiv.table,
#           file = here("tables/fdiv_bioregions_to_annotate.csv"), row.names = FALSE)
```


# 6. Extra: Cross correlation of diveristy metrics
```{r, eval=FALSE}
fundiv.ind.biom <- fundiv.biom$functional_diversity_indices %>% 
  rownames_to_column("Key") %>% 
  select(Key, Sp.Ric = sp_richn, FRic = fric, FDis = fdis, FEve = feve, 
         FDiv = fdiv, FOri = fori)  %>% 
  left_join(select(zoopmeta, Key, Bioregion), by = "Key") %>% 
  relocate(Bioregion) %>% 
  column_to_rownames("Key")

colnames(fundiv.ind.biom) <- c("Bioregion","Species Richness","Functional Richness",
                 "Functional Dispersion","Functional Evenness",
                 "Functional Divergence","Functional Originality")


# Cross correlate metrics
library(GGally)
gp <- GGally::ggpairs(fundiv.ind.biom[,-1], 
                      aes(color = fundiv.ind.biom$Bioregion, alpha = 0.2),
                      lower = list(continuous = "smooth"), progress = FALSE) 
gp +
  scale_color_manual(values = clrs.bioregion) +
  scale_fill_manual(values = clrs.bioregion)

```






