---
title: "Functional diversity maps"
author: "Patrick Pata"
date: "2024-06-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load libraries and data
```{r}
library(tidyverse)
library(rnaturalearth)
library(marmap)
library(mapdata)
library(cowplot)
library(here)
library(ggtext)

theme_set(theme_bw())

# Load data
load(here("data/zoop_data_regionalized_annotated_20231018.RData"))

# The calculated diversity metrics for 163 species
load(here("data/fundiv_3721_163species_20240618.RData"))
```


# 2. Load results from file 8-Functional diversity indices.Rmd
Using the relative abundance or biomass to calculate diversity produces similar results only for species richness and functional richness.
```{r}
fundiv.ind.biom <- fundiv.biom$functional_diversity_indices %>% 
  rownames_to_column("Key") %>% 
  select(Key, Sp.Ric = sp_richn, FRic = fric, FDis = fdis, FEve = feve, 
         FDiv = fdiv, FOri = fori)  %>% 
  left_join(select(zoopmeta, Key, Bioregion), by = "Key") %>% 
  relocate(Bioregion) %>% 
  column_to_rownames("Key")

colnames(fundiv.ind.biom) <- c("Bioregion","Species Richness","Functional Richness",
                 "Functional Dispersion","Functional Evenness",
                 "Functional Divergence","Functional Originality")


# Cross correlate metrics
library(GGally)
gp <- GGally::ggpairs(fundiv.ind.biom[,-1], 
                      aes(color = fundiv.ind.biom$Bioregion, alpha = 0.2),
                      lower = list(continuous = "smooth"), progress = FALSE) 
gp +
  scale_color_manual(values = clrs.bioregion) +
  scale_fill_manual(values = clrs.bioregion)


ggsave(filename = here("figures/fig_s8_fdiversity_bioregions_biomass.png"),
    width = 15, height= 10, units = "in", dpi = 300, bg = "white")
```

# 3. Violin plots
```{r}
Metric.labs <- c("Species Richness","Functional Richness",
                 "Functional Dispersion","Functional Evenness",
                 "Functional Divergence","Functional Originality")
names(Metric.labs) <- c("Sp.Ric","FRic","FDis","FEve","FDiv","FOri")


fdivplots <- vector("list",2)
count <- 1
for (i in c(1:6)) {
   fdivplots[[count]] <- ggplot(filter(fundiv.ind.biom.long, 
                                        Metric == names(Metric.labs)[i]),
                      aes(x = Bioregion, y = Value, fill = Bioregion)) +
  geom_violin(aes(color = Bioregion), alpha = 0.6) + 
  geom_boxplot(width = 0.3, color = "grey25", outlier.shape = NA) +
  xlab("") +
  ylab(Metric.labs[i]) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme(legend.position = "none",  
        axis.title = element_text(size = 12),
        axis.text.x = element_blank(),
        axis.text.y = element_text(size = 12),
        axis.ticks.x = element_blank(),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        strip.text.x = element_text(size = 12)) +
  scale_fill_manual(values = clrs.bioregion) +
  scale_color_manual(values = clrs.bioregion) 
  
  count <- count + 1
}

legend_b1 <- get_legend(
   fdivplots[[1]] + 
    # guides(color = guide_legend(nrow = 2)) +
    theme(legend.position = "bottom")
)


plot.violin <- plot_grid(
  plot_grid(plotlist = fdivplots, nrow = 1, labels = c("c","d","e","f","g","h")),
  legend_b1, 
  ncol = 1, rel_heights = c(1, .1))

# ggsave(file = here("figures/fig_7_fundiv_violin.png"),
#        height = 4, width = 16, dpi = 300, bg = "white")
```


# 4. Mapped functional diversity indices
## Base map
```{r}
world <- ne_countries(scale = "large", returnclass = "sf")

# get bathymetry data
bathy = getNOAA.bathy(lon1 = -150, lon2 = -122, lat1 = 45, lat2 = 60, resolution = 1)
# convert bathymetry to data frame
bathy = fortify.bathy(bathy)

basemap <- ggplot(data = world) +
  geom_sf(fill = "#A9A9A9") + 
  coord_sf(xlim = c(-145.25, -122), ylim = c(47, 56.1), expand = FALSE) +
  # geom_contour(data = bathy, aes(x=x, y=y, z=z), 
  #              breaks=c(-100,-200,-500), linewidth=0.2, colour="black") +
  xlab("") + ylab("") + theme_bw() + 
  theme(legend.position = "bottom", 
        legend.key.width = unit(0.8, "inch"),
        # legend.key.height = unit(0.5, "inch"),
        legend.title = element_text(size = 14),
        legend.text = element_text(size = 12), 
        plot.title = element_text(size = 12, face = "bold"),
        axis.title = element_text(size = 12), 
        axis.text = element_text(size = 12)) +
  scale_x_continuous(breaks = c(-145,-140,-135,-130,-125))

# save(bathy, basemap, file = here("data/backup_bathy_getNOAA_file.RData"))
```


## Calculate functional diversity values in grid cells
```{r, fig.width=10, fig.height=7}
res <- 0.2
lon_grid <- seq(-145.25,-122,res)
lat_grid <- seq(47, 56, res)


bioregions.gridded <- zoopmeta %>%
  select(Key,Bioregion,Longitude,Latitude) %>%
  mutate(lon_bin = lon_grid[ .bincode(Longitude, breaks = lon_grid) ],
         lat_bin = lat_grid[ .bincode(Latitude, breaks = lat_grid) ]) %>%
  group_by(Bioregion, lon_bin, lat_bin) %>%
  summarise(nsamples = n()) %>%
  group_by(lon_bin, lat_bin) %>%
  top_n(n=1) %>%
  arrange(Bioregion)
# There are duplicated grid points if grouped by bioregion
bioregions.gridded$isdistinct <- duplicated(select(bioregions.gridded, lon_bin, lat_bin))


fundiv.grid <- fundiv.ind.biom.long %>% 
  left_join(select(zoopmeta, Key, Longitude, Latitude),
            by = "Key") %>% 
  mutate(lon_bin = lon_grid[ .bincode(Longitude, breaks = lon_grid) ], 
         lat_bin = lat_grid[ .bincode(Latitude, breaks = lat_grid) ]) %>% 
  group_by(Metric, lon_bin, lat_bin) %>%  # TODO check if grouping by bioregion
  summarise(nsamples = n(),
            Metric.Mean = mean(Value),
            Metric.SD = sd(Value),
            Metric.Min = min(Value),
            Metric.Max = max(Value),
            .groups = "drop")

# Individually plot the metrics
div.nsamp <- basemap +
  geom_tile(data = filter(fundiv.grid, 
                          Metric == "Sp.Ric"),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=nsamples),
            alpha = 0.8) +
  scale_fill_viridis_c(name = "Number of samples",
                       trans = "log10",
                       breaks = c(1,3,10,30,80)) +
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200, -500), size=c(0.2), colour="black")

# ggsave(file = here("figures/div_map_N_samples.png"),
#        height = 6, width = 8, dpi = 300)

div.spric <- basemap +
  geom_tile(data = filter(fundiv.grid, 
                          Metric == "Sp.Ric"),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=Metric.Mean),
            alpha = 0.8) +
  scale_fill_viridis_c(name = "Species Richness") +
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200, -500), size=c(0.2), colour="black") +
  ggtitle("Species Richness")

# ggsave(file = here("figures/div_map_species_richness.png"),
#        height = 6, width = 8, dpi = 300)



div.fric <- basemap +
  geom_tile(data = filter(fundiv.grid, 
                          Metric == "FRic"),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=Metric.Mean),
            alpha = 0.8) +
  scale_fill_viridis_c(name = "Functional Richness",
                       trans = "log10",
                       breaks = c(0.02,0.05,0.1,0.2,0.4,0.6)) +
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200, -500), size=c(0.2), colour="black")

# ggsave(file = here("figures/div_map_functional_richness.png"),
#        height = 6, width = 8, dpi = 300)


div.fdis <- basemap +
  geom_tile(data = filter(fundiv.grid, 
                          Metric == "FDis"),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=Metric.Mean),
            alpha = 0.8) +
  scale_fill_viridis_c(name = "Functional Dispersion",
                       breaks = c(0.1,0.2,0.3,0.4,0.5,0.6,0.7)) +
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200, -500), size=c(0.2), colour="black")
# ggsave(file = here("figures/div_map_functional_dispersion.png"),
#        height = 6, width = 8, dpi = 300)


div.feve <- basemap +
  geom_tile(data = filter(fundiv.grid, 
                          Metric == "FEve"),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=Metric.Mean),
            alpha = 0.8) +
  scale_fill_viridis_c(name = "Functional Evenness",
                       # trans = "log10",
                       breaks = c(0.1,0.2,0.3,0.4,0.5,0.6,0.7)) +
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200, -500), size=c(0.2), colour="black")

# ggsave(file = here("figures/div_map_functional_evenness.png"),
#        height = 6, width = 8, dpi = 300)


div.fdiv <- basemap +
  geom_tile(data = filter(fundiv.grid, 
                          Metric == "FDiv"),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=Metric.Mean),
            alpha = 0.8) +
  scale_fill_viridis_c(name = "Functional Divergence",
                       trans = "log10",
                       breaks = c(0.5,0.6,0.7,0.8,0.9)) +
  geom_contour(data = bathy, aes(x=x, y=y, z=z),
               breaks=c(-100,-200, -500), size=c(0.2), colour="black")
# ggsave(file = here("figures/div_map_functional_divergence.png"),
#        height = 6, width = 8, dpi = 300)


div.fori <- basemap +
  geom_tile(data = filter(fundiv.grid, 
                          Metric == "FOri"),
            aes(x=lon_bin+(res/2), y = lat_bin+(res/2),
                width = res, height = res, fill=Metric.Mean),
            alpha = 0.8) +
  scale_fill_viridis_c(name = "Functional Originality",
                       trans = "log10") +
  geom_contour(data = bathy, aes(x=x, y=y, z=z), 
               breaks=c(-100,-200, -500), size=c(0.2), colour="black") 
# ggsave(file = here("figures/div_map_functional_originality.png"),
#        height = 6, width = 8, dpi = 300)
```


# 5. Output merged figures
```{r, fig.width=8}
plot_grid(
  plot_grid(div.fric, div.fdiv,
            nrow = 1, labels = c("a","b")),
  plot.violin,
  ncol = 1,
  rel_heights = c(1.5,1))

ggsave(file = here("figures/fig_7_fundiv_to_annotate.png"),
       height = 10, width = 16, dpi = 300, bg = "white")



plot_grid(div.fdis, div.feve, div.fori,
          align = "h", axis = "bt",
          ncol = 1, labels = c("a","b","c"))

ggsave(file = here("figures/fig_s9_fundiv_maps.png"),
       height = 16, width = 8, dpi = 300, bg = "white")
```

