---
title: "Additional analysis"
author: "Patrick Pata"
date: "2024-10-16"
output: html_document
---

This file contains the analysis of the functional assemblages which are identified based on community level trait characteristics rather than species distributions. The methodology of deriving the functional assemblages mirrors the methodology of deriving the taxonomic bioregions in Pata et al. (2022). Most of the figures and tables here could be reproduced by data outputs from other files except for Fig S8 which requires the restricted raw species-stage resolution community data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Load libraries and data
```{r}
library(tidyverse)
library(cowplot)
library(here)
library(ggtext)
library(vegan)
library(ggrepel)
theme_set(theme_bw())
set_here(path = "..")
`%notin%` <- Negate(`%in%`)

load(here("data/fundiv_3721_163species_20240618.RData"))
load(here("data/community_metrics_3721_163species_20250217.RData")) 
load(here("data/bc_bioregionalization_pca_20250217.RData"))
load(here("data/BC_bioregionalization_metadata.RData"))

geomean <- function(x) {  exp(mean(log(x), na.rm = TRUE))  }
`%notin%` <- Negate(`%in%`)
source(here("Functions/p_biplot_pca.R"))
source(here("Functions/p_biplot_pca_points.R"))
source(here("functions/p_goodness_of_clustering_v2.R"))


fungrp.colors <- c("turquoise3", "steelblue2","darkorange2",
                   "orchid","magenta4","lightcoral",
                   "gold3","darkgreen","darkolivegreen4",
                   "olivedrab3")
clrs.bioregion <- c("#004DFB","#FB20CA","#FFA500","#20FF00")
```


# 2. Organize the functional characteristics, log transform some values
```{r}
# Select total respiration rate and clearance rate because these are least similar, although r = 0.85
ctotal <- zoop.163.ctotal %>% 
  filter(Trait %in% c("totalRespirationRate","totalClearanceRate")) %>% 
  # Log transform the values then widen
  mutate(CTotal = log(CTotal)) %>% 
  pivot_wider(names_from = "Trait", values_from = "CTotal") %>% 
  select(Key, Bioregion, totalRespirationRate, totalClearanceRate)

# The CWMs in long format are in zoop.163.cwm, convert to wider format for the figure
# Reorder according to pca.zoop order
cwm.wide <- data.frame(Key = rownames(fundiv.ind.biom)) %>% 
  left_join(zoop.163.cwm, by = "Key", multiple = "all") %>% 
  pivot_wider(names_from = Trait, values_from = CWM) %>% 
  column_to_rownames("Key") %>% 
  # Log transform specific traits (this will not affect correlations but will affect the PCA of the cWMs)
  mutate(bodyLength = log(bodyLength),
         dryWeight = log(dryWeight),
         carbonWeight = log(carbonWeight),
         nitrogenWeight = log(nitrogenWeight),
         phosphorusWeight = log(phosphorusWeight),
         respirationRate = log(respirationRate),
         clearanceRate = log(clearanceRate),
         excretionRateN = log(excretionRateN),
         excretionRateP = log(excretionRateP),
         respirationRate_WSC = log(respirationRate_WSC),
         clearanceRate_WSC = log(clearanceRate_WSC),
         excretionRateN_WSDW = log(excretionRateN_WSDW),
         excretionRateP_WSDW = log(excretionRateP_WSDW))

for.pca <- cwm.wide %>% 
  rownames_to_column("Key") %>% 
  left_join(ctotal, 
            by = join_by(Key, Bioregion)) %>% 
  left_join(fundiv.ind.biom %>% 
              rownames_to_column("Key"),
            by = join_by(Key, Bioregion)) %>% 
  column_to_rownames("Key") %>% 
  dplyr::select(-Bioregion) %>% 
  
  # Rename columns
  rename("Active ambush" = "active ambush", "Cruise" = "cruise",
         "Passive ambush" = "passive ambush", "Current" = "current",
         "Broadcasting" = "broadcasting", "Brooding" = "brooding",
         "Carnivore" = "carnivore", "Detritivore" = "detritivore",
         "Herbivore" = "herbivore", "Omnivore" = "omnivore",
         "Body length" = "bodyLength", 
         "%C" = "carbonPDW",
         "C weight" = "carbonWeight", 
         "Ind. clearance" = "clearanceRate",
         "Ws. clearance" = "clearanceRate_WSC", 
         "Dry weight" = "dryWeight",
         "Ind. N excretion" = "excretionRateN",
         "Ws. N excretion" = "excretionRateN_WSDW",
         "Ind. P excretion" = "excretionRateP",
         "Ws. P excretion" = "excretionRateP_WSDW",
         "%water" = "waterPWW",
         "%N" = "nitrogenPDW", 
         "N weight" = "nitrogenWeight",
         "%P" = "phosphorusPDW", 
         "P weight" = "phosphorusWeight",
         "C:N" = "ratioCN", "C:P" = "ratioCP", "N:P" = "ratioNP",
         "Ind. respiration" = "respirationRate",
         "Ws. respiration" = "respirationRate_WSC",
         "Total respiration" = "totalRespirationRate",
         "Total clearance" = "totalClearanceRate",
         "Species richness" = "Sp.Ric", "Functional richness" = "FRic",
         "Functional dispersion" = "FDis", "Functional evenness" = "FEve",
         "Functional divergence" = "FDiv", "Functional originality" = "FOri")
  
```


# 3. PCAs
## PCA of original bioregionalization biplot with functional characteristics
```{r}
# # Uncomment this to load the original bioregionalization site x species matrix for the PCA used in Pata et al. (2022). The PCA object is already loaded above.
# load(here("data/restricted_data/zoop_data_regionalized_10252021.RData"))
# rm(list=setdiff(ls(), ""))
# pca.zoop <- rda(zoop.logchord)
# save(pca.zoop, file = here("data/bc_bioregionalization_pca_20250217.RData"))


# Plot the PCA of the species distribution used in the taxonomic bioregions
# Correlate the axes 
axismat <- vegan::scores(pca.zoop, choices = c(1:4), display = "sites")
points.stand <- scale(axismat)
xmat = scale(for.pca , center=TRUE, scale=TRUE)

U <- cor(xmat, points.stand, use="pairwise.complete.obs")
# Only retain correlations > 0.25 for axes 1 & 2
U12 <- as.data.frame(U) %>%
  filter( abs(PC1) > 0.25 | abs(PC2) > 0.25 )

# Only retain correlations > 0.25 for axes 3 & 4
U34 <- as.data.frame(U) %>%
  filter( abs(PC3) > 0.25 | abs(PC4) > 0.25 )

# Filter the relatively strong correlations
for.pca12 <- as.data.frame(for.pca) %>% 
  dplyr::select(rownames(U12))
for.pca34 <- as.data.frame(for.pca) %>%  
   dplyr::select(rownames(U34))

p1a <- p_biplot_pca_points(rdaoutput = pca.zoop, 
             xmat = for.pca12,
             cluster = cwm.wide$Bioregion,
             axestoplot = c(1,2),
             alpha = 0.1, legendtitle = "Bioregion",
             clrs = clrs.bioregion, 
             rescaleSites = 2, scaleCorrelations = FALSE) +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))


p1b <- p_biplot_pca_points(rdaoutput = pca.zoop, 
             xmat = for.pca34,
             cluster = cwm.wide$Bioregion,
             axestoplot = c(3,4),
             alpha = 0.1, legendtitle = "Bioregion",
             clrs = clrs.bioregion, 
             rescaleSites = 2, scaleCorrelations = FALSE) +
  theme(legend.position = "bottom")+
  guides(color = guide_legend(override.aes = list(alpha = 1)))
```


## Check correlation between community level traits
```{r,eval=FALSE}
# Find strongest correlations
# cor.mat <- cor(cwm.wide[,2:31], use = "complete.obs")
cor.mat <- cor(for.pca, use = "complete.obs")
diag(cor.mat) <-  NA
cor.mat.long <- cor.mat %>% 
  as.data.frame() %>% 
  rownames_to_column("IndexY") %>% 
  pivot_longer(cols = -IndexY, names_to = "IndexX", values_to = "Value") %>% 
  filter(!is.na(Value)) %>% 
  mutate(IndexX = factor(IndexX, levels = colnames(cor.mat)),
         IndexY = factor(IndexY, levels = colnames(cor.mat)),)
cor.strong <- cor.mat.long %>% 
  filter(abs(Value) > 0.66) %>% 
  arrange(-abs(Value))
cor.strong <- cor.strong[seq(1,nrow(cor.strong),2),]
```


## PCA of community level trait values
```{r}
# Exclude traits that are strongly correlated 
for.pca.sub <- for.pca %>% 
  dplyr::select(-c("Ind. respiration", "N weight", "P weight",
            "Dry weight","Ind. P excretion", 
            "Ind. N excretion", "Ind. clearance",
            "%C","%P", "%N", "Broadcasting",
            "Ws. P excretion",
            "C:P","Omnivore","Carnivore","Species richness",
            "Total clearance","Body length"))
  
for.pca.sub <- scale(for.pca.sub)

pca.ctv <- rda(for.pca.sub)


# visualize all community level traits instead of the selected 20?
# Correlate the axes
axismat <- vegan::scores(pca.ctv, choices = c(1:4), display = "sites")
points.stand <- scale(axismat)
xmat = scale(for.pca , center=TRUE, scale=TRUE)

U <- cor(xmat, points.stand, use="pairwise.complete.obs")
# Only retain correlations > 0.25 for axes 1 & 2
U12 <- as.data.frame(U) %>%
  filter( abs(PC1) > 0.25 | abs(PC2) > 0.25 )

# Only retain correlations > 0.25 for axes 3 & 4
U34 <- as.data.frame(U) %>%
  filter( abs(PC3) > 0.25 | abs(PC4) > 0.25 )

# Filter the relatively strong correlations
for.pca12 <- as.data.frame(for.pca) %>%
  dplyr::select(rownames(U12))
for.pca34 <- as.data.frame(for.pca) %>%
   dplyr::select(rownames(U34))


# Variance explained
head(round( pca.ctv$CA$eig/pca.ctv$tot.chi*100, digits = 1),10)
sum(head(round( pca.ctv$CA$eig/pca.ctv$tot.chi*100, digits = 1),6))


p2a <- p_biplot_pca_points(rdaoutput = pca.ctv, 
             xmat = for.pca12, #for.pca.sub,
             cluster = cwm.wide$Bioregion,
             axestoplot = c(1,2),
             alpha = 0.1, legendtitle = "Bioregion",
             clrs = clrs.bioregion, 
             rescaleSites = 0.7, scaleCorrelations = FALSE) +
  scale_y_reverse() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

p2b <- p_biplot_pca_points(rdaoutput = pca.ctv, 
             xmat = for.pca34, #for.pca.sub,
             cluster = cwm.wide$Bioregion,
             axestoplot = c(3,4),
             alpha = 0.1, legendtitle = "Bioregion",
             clrs = clrs.bioregion, 
             rescaleSites = 0.7, scaleCorrelations = FALSE)+
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))
```

# 4. Cluster analysis
Cluster the CTVs to get clusters with similarities in functional characteristics
With 5 axes representing 68% of the variance in the data, 6 clusters were optimal using the K-means analysis.

Reassign clusters
2 -> FA1 (current feeding, high biomass, copepod/appendicularian)
3 -> FA2 (Deepshelf + ~OS, euphausiid dominated)
4 -> FA3 (Nearshore + ~DS, coastal gelatinous dominated)
1 -> FA4 (deepfjord + offshore... deep and diverse with FG4/detri & gelat)
5 -> FA5 (Offshore + deepshelf, "balanced", high func disp, more chaetognaths)
6 -> FA6 (outliers)
```{r}
# The first 5 axes contain 68% of the variance (> 2/3)
sum(head(round(pca.ctv$CA$eig/pca.ctv$tot.chi*100, digits = 1),5))
pca.ctv.scores <- vegan::scores(pca.ctv, choices = c(1:5))$sites

# Hierarchical
pca.hc <- hclust(dist(pca.ctv.scores), method = "ward.D2")
plot(pca.hc, labels = FALSE)

set.seed(888)
nclust <- 15
hclust.funreg<- matrix(NA, nrow = nrow(pca.ctv.scores), ncol = (nclust-1))
kmeans.funreg<- matrix(NA, nrow = nrow(pca.ctv.scores), ncol = (nclust-1))
for (k in seq(2,nclust)) {
  hclust.funreg[,k-1] <- as.factor(cutree(pca.hc, k = k))
  kmeans.funreg[,k-1] <- kmeans(pca.ctv.scores, k, nstart = 10, 
                                iter.max = 1000)$cluster
}
# gci.h <- goc_index(multiclust = hclust.funreg,
#                 distmat = dist(pca.ctv.scores)) %>% 
#   pivot_longer(-Num_Clusters, names_to = "Index", 
#                values_to = "Index Value") %>% 
#   mutate(method = "ward")
# gci.k <- goc_index(multiclust = kmeans.funreg,
#                 distmat = dist(pca.ctv.scores)) %>% 
#   pivot_longer(-Num_Clusters, names_to = "Index", 
#                values_to = "Index Value") %>% 
#   mutate(method = "kmeans")
# 
# gci.both <- bind_rows(gci.h, gci.k)
# 
# ggplot(gci.both, aes(x = Num_Clusters, y = `Index Value`, color = method)) + 
#   geom_point() + geom_line() +
#   scale_x_continuous(breaks = seq(2,20,1)) +
#   xlab("Number of Clusters") +
#   facet_wrap(~ Index, nrow = 2, scales = "free")
#
#
# ggsave(file = here("figures/gci_ctv.png"),
#        height = 6, width = 8, dpi = 300, bg = "white")


# Appears that 6 is best
pca.meta <- as.data.frame(pca.ctv.scores) %>% 
  rownames_to_column("Key") %>% 
  dplyr::select(Key) %>% 
  mutate(FunRegion = as.character(kmeans.funreg[,5])) %>% 
  # rename and rearrange cluster ids
  mutate(FunRegion = if_else(FunRegion == "1", "FA4", FunRegion),
         FunRegion = if_else(FunRegion == "2", "FA1", FunRegion),
         FunRegion = if_else(FunRegion == "3", "FA2", FunRegion),
         FunRegion = if_else(FunRegion == "4", "FA3", FunRegion),
         FunRegion = if_else(FunRegion == "5", "FA5", FunRegion),
         FunRegion = if_else(FunRegion == "6", "FA6", FunRegion)) %>% 
  mutate(FunRegion = factor(FunRegion,
                            levels = c("FA1","FA2","FA3","FA4","FA5","FA6")))


clrs.funbioregion2 <- c("gold","slateblue2","firebrick1","turquoise3","sienna2","gray65")


p3a <- p_biplot_pca(rdaoutput = pca.ctv, 
             # xmat = for.pca,
             cluster = pca.meta$FunRegion,
             axestoplot = c(1,2),
             alpha = 0.4, legendtitle = "Functional\nAssemblage",
             clrs = clrs.funbioregion2, plottitle = "") +
  scale_y_reverse() +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

p3b <- p_biplot_pca(rdaoutput = pca.ctv, 
             # xmat = for.pca,
             cluster = pca.meta$FunRegion,
             axestoplot = c(3,4),
             alpha = 0.4, legendtitle = "Functional\nAssemblage",
             clrs = clrs.funbioregion2, plottitle = "")  +
  theme(legend.position = "bottom") +
  guides(color = guide_legend(override.aes = list(alpha = 1)))

# # Output these clusters for further analysis
# save(pca.meta, clrs.funbioregion2, gci.both, kmeans.funreg,
#      file = here("data/clustering_fun_assemblages.RData"))

```



# 5. Correlate the species biomass to the CWM-PCA axes
Color by functional group. This requires the species distribution data.
```{r}
# Community data
load(here("data/restricted_data/key_fg_abundance_biomass_20240626.RData"))

# Organize the site x biomass matrix
com.mat <- zoop.bc.species.fun %>% 
  # If based on biomass
  dplyr::select(c(Key, Species, Biomass.mg.m2)) %>% 
  pivot_wider(names_from = "Species", values_from = "Biomass.mg.m2",
              values_fill = 0) %>%
  column_to_rownames("Key")
  
# Transform the matrix
com.mat.2 <- log1p(com.mat)

# PCA axis scores
pca.ctv.scores <- vegan::scores(pca.ctv, choices = c(1:5))$sites

# Correlate species abundances to PCA axis scores
U <- cor(com.mat.2, pca.ctv.scores) %>% 
  as.data.frame() %>% 
  rownames_to_column("Species") %>% 
  # If by functional group
  left_join(zoop.bc.species.fun %>% 
              distinct(Species, fungrp), 
            by = "Species") %>% 
  mutate(fungrp = factor(fungrp, levels = c(1:10)))

U1 <- filter(U, PC1 >0.20 | PC1 < -0.20 | PC2 < -0.25 | PC2 > 0.15)
p4a <- ggplot(U, aes(x = PC1, y = PC2)) +
  geom_vline(xintercept = 0, colour='darkgray') + 
  geom_hline(yintercept = 0, colour='darkgray') +
  geom_point(shape = 16, size = 2, aes(color = fungrp)) +
  geom_text_repel(data = U1,
                  aes(label = Species, color = fungrp), size = 4,
                  max.overlaps = 15, fontface = "italic",
                  show.legend = FALSE) +
  scale_y_reverse() +
  scale_color_manual(values = rev(fungrp.colors[sort(as.numeric(unique(U$fungrp)))]),
                     name = "Functional\nGroup",
           labels = paste0("FG", unique(zoop.163.ctotal.ave$fungrp)),
                     guide = guide_legend(override.aes = list(size = 4))) +
  theme(aspect.ratio = 1, legend.position = "bottom",
        legend.title = element_text(color = 'black', size = 12, face = 'bold'),
        legend.text = element_text(size = 12))

U2 <- filter(U, PC3 > 0.45 | PC3 < -0.2 | PC4 > 0.20 | PC4 < -0.25)
p4b <- ggplot(U, aes(x = PC3, y = PC4)) +
  geom_vline(xintercept = 0, colour='darkgray') + 
  geom_hline(yintercept = 0, colour='darkgray') +
  geom_point(shape = 16, size = 2, aes(color = fungrp)) +
  geom_text_repel(data = U2,
                  aes(label = Species, color = fungrp), size = 4,
                  max.overlaps = 15, fontface = "italic",
                  show.legend = FALSE) +
  scale_color_manual(values = rev(fungrp.colors[sort(as.numeric(unique(U$fungrp)))]),
                     name = "Functional\nGroup",
           labels = paste0("FG", unique(zoop.163.ctotal.ave$fungrp)),
                     guide = guide_legend(override.aes = list(size = 4))) +
  theme(aspect.ratio = 1, legend.position = "bottom",
        legend.title = element_text(color = 'black', size = 12, face = 'bold'),
        legend.text = element_text(size = 12))  


```



# 6. Fig5 FigS7 Combine ordinations in one figure
```{r, fig.height=6, fig.width=6}
cowplot::plot_grid(p1a,p2a,p3a,p4a,
                   nrow = 2,
                   labels = "auto")
ggsave(filename = here("figures/fig_6_FA_ordinations_12_Aug2025.png"),
       width = 12, height= 13, units = "in", dpi = 600, bg = "white")

# Supplementary figure for axes 3 & 4
cowplot::plot_grid(p1b,p2b,p3b,p4b,
                   nrow = 2,
                   labels = "auto")
ggsave(filename = here("figures/fig_S5_FA_ordinations_34_Aug2025.png"),
       width = 12, height= 13, units = "in", dpi = 600, bg = "white")


```



# 7. FigS6 Correspondence FA and bioregions
## Time series of the FC clusters by bioregion
Biased by number of samples in a season-year
```{r}
# By year
ts.fclust <- pca.meta %>% 
  left_join(zoopmeta, by = "Key") %>% 
  group_by(Bioregion, Year, FunRegion) %>% 
  summarise(nsamples = n(), .groups = "drop")

ggplot(ts.fclust, aes(x = Year, y = nsamples)) +
  geom_bar(aes(fill = FunRegion), position = "fill", stat="identity") +
  scale_fill_manual(values = clrs.funbioregion2,
           name = "Functional\nAssemblage") +
  scale_y_continuous(expand = c(0, 0),
                     name = "Proportion of samples") +
  theme(legend.position = "right",  
        strip.text = element_text(size = 12),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        axis.title.y.right = element_markdown()) +
  facet_wrap(~Bioregion, scales = "free_y")

# ggsave(filename = here("figures/fun_assem_timeseries_year.png"),
#        width = 9, height= 6, units = "in", dpi = 600, bg = "white")


# By month
ts.fclust <- pca.meta %>% 
  left_join(zoopmeta, by = "Key") %>% 
  group_by(Bioregion, Month, FunRegion) %>% 
  summarise(nsamples = n(), .groups = "drop")

pfa1 <- ggplot(ts.fclust, aes(x = as.factor(Month), y = nsamples)) +
  geom_bar(aes(fill = FunRegion), position = "fill", stat="identity") +
  scale_fill_manual(values = clrs.funbioregion2,
           name = "Functional\nAssemblage") +
  scale_y_continuous(expand = c(0, 0),
                     name = "Proportion of samples") +
  xlab("Month") +
  theme(legend.position = "right",  
        strip.text = element_text(size = 12),
        axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        axis.title.y.right = element_markdown()) +
  facet_wrap(~Bioregion, scales = "free_y")


pfa1

# ggsave(filename = here("figures/fun_assem_timeseries_month.png"),
#        width = 9, height= 6, units = "in", dpi = 600, bg = "white")
```

## Table matching the original bioregionalization with the functional clusters
```{r}
confmat <- table(zoopmeta$Bioregion, pca.meta$FunRegion) %>% 
  as.data.frame()
colnames(confmat) = c("Bioregion","FunCluster","Nsamples")
# Calculate the percent of samples from the bioregion belonging to the functional clusters
confmat <- confmat %>% 
  group_by(Bioregion) %>% 
  mutate(totalsamp = sum(Nsamples)) %>% 
  ungroup() %>% 
  mutate(perc.bioregion = Nsamples/totalsamp*100) %>% 
  
  group_by(FunCluster) %>% 
  mutate(totalFC = sum(Nsamples)) %>% 
  ungroup() %>% 
  mutate(perc.FC = Nsamples/totalFC*100)


pfa2 <- ggplot(confmat, aes(x = FunCluster, y = Bioregion)) +
  geom_tile(aes(fill = Bioregion, alpha = perc.bioregion), color = "white") + 
  geom_text(aes(label = paste0(sprintf("%0.1f", perc.bioregion), "%\n (n = ",
                               Nsamples,")")), 
            color = "black") +
  scale_fill_manual(values = clrs.bioregion) +
  xlab("Functional Assemblage") +
  theme(axis.title = element_text(size = 12),
        axis.text.x = element_text(size = 12, 
                                   face = "bold", colour = clrs.funbioregion2),
        axis.text.y = element_text(size = 12, 
                                   face = "bold", colour = clrs.bioregion),
        legend.position = "none",
        panel.background = element_rect(fill = "white"),
         panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) 
pfa2


# Combine the overall and seasonal matching between bioregions
cowplot::plot_grid(pfa2, pfa1, labels = "auto",
                   rel_heights = c(1, 1.8), ncol = 1)

ggsave(filename = here("figures/fig_s6_fun_assem_bioregion.png"),
       width = 9, height= 8, units = "in", dpi = 600, bg = "white")

```



# 8. Fig S8 Total rates figure but by functional clusters
```{r, eval=FALSE}
# Load the complete community data for community totals
load(here("data/restricted_data/BC_3721_163_speciesstage_traits_20250217.RData"))


zoop.163.ctotal.plot.2 <- zoop.spstg.ctotal %>%
  left_join(pca.meta, by = "Key") %>% 
  
  mutate(Abundance.m2 = `Abundance(#/m3)` * Depth_Start) %>% 
  mutate(Biomass.mg.m2 = `Biomass(mg/m3)` * Depth_Start) %>% 

  # Calculate community totals
  group_by(Key, FunRegion, fungrp) %>%
  summarise(dryWeight = sum(dryWeight * Abundance.m2),
            carbonWeight = sum(carbonWeight * Abundance.m2),
            nitrogenWeight = sum(nitrogenWeight * Abundance.m2),
            phosphorusWeight = sum(phosphorusWeight * Abundance.m2),
            respirationRate = sum(respirationRate * Abundance.m2),
            clearanceRate = sum(clearanceRate * Abundance.m2),
            excretionRateN = sum(excretionRateN * Abundance.m2),
            excretionRateP = sum(excretionRateP * Abundance.m2),
            Abundance = sum(Abundance.m2),
            Biomass = sum(Biomass.mg.m2),
            .groups = "drop") %>% 
  
  # pad zeros to empty FGs (not really necessary... because the Key-Bioregion-FG pair with blanks will never show up (e.g., FG2))
  pivot_longer(cols = c(dryWeight:Biomass),
               names_to = "Metric", values_to = "Value") %>% 
  pivot_wider(names_from = Metric, values_from = Value,
              values_fill = 0) %>%
  pivot_longer(cols = c(dryWeight:Biomass),
               names_to = "Metric", values_to = "Value") %>% 
  
  # Average across all values (not staggered by year-month)
  group_by(FunRegion, Metric, fungrp) %>% 
  summarise(Value.mean = mean(Value),
            Value.sd = sd(Value),
            nsamples = n(), .groups = "drop") %>% 
  mutate(Value.se = Value.sd / sqrt(nsamples)) %>% 
  mutate(fungrp = factor(fungrp, levels = c("1","2","3","4","5","6",
                                            "7","8","9","10"))) %>%  
  
  filter(Metric %notin% c("Biomass","nitrogenWeight", 
                          "phosphorusWeight", "dryWeight")) %>% 
  mutate(Metric = if_else(Metric == "carbonWeight",
                          "Carbon Weight", Metric),
         Metric = if_else(Metric == "clearanceRate",
                          "Clearance Rate", Metric),
         Metric = if_else(Metric == "respirationRate",
                          "Respiration Rate", Metric),
         Metric = if_else(Metric == "excretionRateN",
                          "Ammonium Excretion Rate", Metric),
         Metric = if_else(Metric == "excretionRateP",
                          "Phosphate Excretion Rate", Metric)) %>% 
  mutate(Metric = factor(Metric, levels = c("Abundance","Carbon Weight",
                                            "Respiration Rate",
                                            "Ammonium Excretion Rate",
                                            "Phosphate Excretion Rate",
                                            "Clearance Rate")))
           

# Cowplot version
metrics <- levels(zoop.163.ctotal.plot.2$Metric)
plots <- vector("list",2)
count <- 1
for (i in c(1:6)) {
   plots[[count]] <- ggplot(zoop.163.ctotal.plot.2 %>% 
                                  filter(Metric == metrics[i]), 
                                aes(x = FunRegion, y = Value.mean)) +
      geom_bar(aes(fill = fungrp), position = "stack", stat = "identity") +
      scale_fill_manual(values = rev(fungrp.colors), 
                        name = "Functional\nGroup") +
      scale_y_continuous(expand = c(0, 0),
                         name =unique(metrics[i])) +
     xlab("Functional Assemblage") +
      theme(legend.position = "none",
            axis.title.y.right = element_markdown()) 
  
  count <- count + 1
}

# Combine the plots into a grid
plot_grid_layout <- plot_grid(plotlist = plots, ncol = 3, labels = "auto") 
plot_grid_layout


ggsave(filename = here("figures/fig_S8_com_total_rates_by_funreg.png"),
       width = 12, height= 6, units = "in", dpi = 600, bg = "white")
```

# 9. Fig S9 Functional diversity by FA
```{r}
fd.fa <- fundiv.ind.biom.long %>% 
  left_join(pca.meta, by = "Key") 

Metric.labs <- c("Species Richness","Functional Richness",
                 "Functional Dispersion","Functional Evenness",
                 "Functional Divergence","Functional Originality")
names(Metric.labs) <- c("Sp.Ric","FRic","FDis","FEve","FDiv","FOri")


fdivplots <- vector("list",2)
count <- 1
for (i in c(1:6)) {
   fdivplots[[count]] <- ggplot(filter(fd.fa, 
                                        Metric == names(Metric.labs)[i]),
                      aes(x = FunRegion, y = Value, fill = FunRegion)) +
  geom_violin(aes(color = FunRegion), alpha = 0.6) + 
  geom_boxplot(width = 0.3, color = "grey25", outlier.shape = NA) +
  xlab("") +
  ylab(Metric.labs[i]) +
  scale_y_continuous(expand = c(0.01, 0.01)) +
  theme(legend.position = "none",  
        axis.title = element_text(size = 12),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        # axis.ticks.x = element_blank(),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 12),
        strip.text.x = element_text(size = 12)) +
  scale_fill_manual(values = clrs.funbioregion2) +
  scale_color_manual(values = clrs.funbioregion2) 
  
  count <- count + 1
}

legend_b1 <- get_legend(
   fdivplots[[1]] + 
    # guides(color = guide_legend(nrow = 2)) +
    theme(legend.position = "bottom")
)


plot.violin <- plot_grid(
  plot_grid(plotlist = fdivplots, nrow = 2, labels = "auto"),
  legend_b1, 
  ncol = 1, rel_heights = c(1, .1))

plot.violin

ggsave(file = here("figures/fig_S9_fundiv_funassemblages_violin.png"),
       height = 6, width = 10, dpi = 300, bg = "white")
```